---
title: "R Environment Setup"
subtitle: "Installing R, RStudio, and the required packages for this course"
---

## Required Software

| Software | Version | Download |
|----------|---------|----------|
| **R** | 4.3 or later | [r-project.org](https://www.r-project.org/) |
| **RStudio** (recommended IDE) | Latest | [posit.co/downloads](https://posit.co/download/rstudio-desktop/) |
| **Quarto** (for rendering slides) | 1.8 or later | [quarto.org](https://quarto.org/docs/get-started/) |

::: {.callout-box}
**Note:** The slide decks and this website use `eval: false` for all R code blocks, meaning R code is displayed but not executed during rendering. You do **not** need R installed to view the slides or website. However, to run the workshop exercises interactively, you will need R and the packages below.
:::

## Package Installation

Run this once in your R console to install all packages used across the 10 chapters:

```{r}
#| label: install-all
#| eval: false

install.packages(c(
  "dagitty",   # DAG specification and analysis
  "ggdag",     # DAG visualization with ggplot2
  "bnlearn",   # Bayesian network structure and parameter learning
  "ggplot2",   # Data visualization
  "ivreg",     # Instrumental variable regression (2SLS)
  "tidyr",     # Data reshaping (pivot_longer)
  "gRain"      # Exact inference in Bayesian networks (junction tree)
))
```

## Packages by Chapter

| Package | Purpose | Chapters Used |
|---------|---------|---------------|
| `dagitty` | Define and analyze DAGs, d-separation, adjustment sets | 1, 2, 3, 4, 5, 10 |
| `ggdag` | Visualize DAGs using ggplot2 | 1, 2, 3, 10 |
| `bnlearn` | Bayesian network structure and parameter learning | 4, 5, 7, 10 |
| `ggplot2` | All data visualizations | 4, 5, 6, 7, 8, 9, 10 |
| `gRain` | Exact inference via junction trees (companion to bnlearn) | 4, 5 |
| `ivreg` | Instrumental variable regression (two-stage least squares) | 10 |
| `tidyr` | Data reshaping (`pivot_longer`) | 8 |

## Verification Script

After installing, run this script to verify all packages load correctly:

```{r}
#| label: verify-packages
#| eval: false

# Verification: load each package and print version
packages <- c("dagitty", "ggdag", "bnlearn", "ggplot2", "ivreg", "tidyr", "gRain")

for (pkg in packages) {
  if (requireNamespace(pkg, quietly = TRUE)) {
    cat(sprintf("✓ %s %s\n", pkg, packageVersion(pkg)))
  } else {
    cat(sprintf("✗ %s — NOT INSTALLED\n", pkg))
  }
}
```

Expected output (version numbers may vary):

```
✓ dagitty 0.3-4
✓ ggdag 0.2.12
✓ bnlearn 4.9.4
✓ ggplot2 3.5.1
✓ ivreg 0.6-3
✓ tidyr 1.3.1
✓ gRain 1.4.3
```

## Troubleshooting

### bnlearn installation issues on Windows

`bnlearn` requires compilation. If you encounter errors:

1. Install [Rtools](https://cran.r-project.org/bin/windows/Rtools/) (match the version to your R version)
2. Restart R and try `install.packages("bnlearn")` again

### gRain dependency issues

`gRain` depends on Bioconductor packages. If `install.packages("gRain")` fails:

```{r}
#| label: install-grain
#| eval: false

if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("RBGL")
BiocManager::install("Rgraphviz")
install.packages("gRain")
```

### dagitty on older R versions

If `dagitty` installation fails, ensure your R version is 4.3 or later. The package uses features not available in older versions.

## Consistent MNH Variables

These variable names are used consistently across chapters in the R code workshops:

| Variable | Values | R Name |
|----------|--------|--------|
| Staffing | Adequate / Low | `Staffing` |
| Equipment (CPAP) | Available / Unavailable | `Equipment` or `CPAP` |
| Training | Completed / Not Completed | `Training` |
| Quality of Care | Good / Poor | `Quality` |
| Neonatal Mortality Rate | High (>25/1000) / Low (≤25/1000) | `NMR` |
| Maternal Mortality Ratio | High (>300/100K) / Low (≤300/100K) | `MMR` |
| Coverage | >80% / ≤80% | `Coverage` |
