---
title: "Analysis Techniques for Storytelling with Data"
subtitle: "A hands-on R tutorial: filter, group, summarize, compare — and the questions they answer"
author: "Matthew Kuch"
date: 2026-03-28
format:
  html:
    theme: [cosmo, blog-styles.scss]
    toc: true
    toc-depth: 2
    code-copy: true
    code-fold: false
---

This tutorial walks through the core analysis techniques from the [companion blog post](blog.html) — using R and the tidyverse to demonstrate how four simple data operations (filter, group, summarize, compare) map to four modes of data storytelling (descriptive, diagnostic, prescriptive, predictive).

We'll work with a simulated dataset of measles cases across 10 African countries from 1990 to 2023. The data is built inline — no external files needed. If you want to run the code yourself, grab the standalone script: [`tutorial-analysis.R`](tutorial-analysis.R).

## Setup

```{r}
#| label: setup
#| echo: true
#| message: false

library(tidyverse)
```

## Building the Dataset

Real analysis uses real data. But for teaching purposes, a well-constructed simulation lets us focus on the *techniques* without getting stuck on data-cleaning detours. This dataset mirrors the structure of WHO/UNICEF measles surveillance data: country, year, cases per million, region, and income group.

The key feature: a visible inflection point at **2001**, when the WHO/UNICEF Measles Initiative launched coordinated vaccination campaigns across Africa.

```{r}
#| label: build-dataset
#| echo: true

set.seed(42)

countries <- c("Nigeria", "Ethiopia", "DRC", "Tanzania", "Kenya",
               "Uganda", "Mozambique", "Ghana", "Niger", "Mali")

years <- 1990:2023

measles <- expand_grid(country = countries, year = years) |>
  mutate(
    region = case_when(
      country %in% c("Kenya", "Tanzania", "Uganda", "Ethiopia", "Mozambique") ~ "East Africa",
      country %in% c("Nigeria", "Ghana")                                      ~ "West Africa",
      country %in% c("Niger", "Mali")                                         ~ "Sahel",
      country == "DRC"                                                         ~ "Central Africa"
    ),
    income_group = case_when(
      country %in% c("Kenya", "Ghana", "Nigeria")         ~ "Lower-middle",
      country %in% c("Niger", "Mali", "Mozambique", "DRC",
                      "Ethiopia", "Uganda", "Tanzania")    ~ "Low"
    ),
    baseline = case_when(
      country == "Nigeria" ~ 2200, country == "DRC" ~ 2500,
      country == "Niger" ~ 2400, country == "Mali" ~ 2100,
      country == "Ethiopia" ~ 1800, country == "Tanzania" ~ 1500,
      country == "Kenya" ~ 1200, country == "Uganda" ~ 1600,
      country == "Mozambique" ~ 1700, country == "Ghana" ~ 1100
    ),
    decline_rate = case_when(
      country %in% c("Kenya", "Ghana", "Tanzania")       ~ 0.15,
      country %in% c("Ethiopia", "Uganda", "Mozambique") ~ 0.10,
      country %in% c("Nigeria", "DRC", "Niger", "Mali")  ~ 0.06
    ),
    cases_per_million = if_else(
      year <= 2001,
      baseline + rnorm(n(), 0, 200),
      baseline * exp(-decline_rate * (year - 2001)) + rnorm(n(), 0, 50)
    ),
    cases_per_million = pmax(round(cases_per_million), 10),
    era = if_else(year <= 2001, "Pre-Initiative (1990-2001)", "Post-Initiative (2002-2023)")
  ) |>
  select(country, region, income_group, year, era, cases_per_million)
```

A quick look at the shape of the data:

```{r}
#| label: glimpse-data
#| echo: true

glimpse(measles)
```

340 rows: 10 countries x 34 years. Each row has a country, its region, income classification, year, era label, and the number of measles cases per million people.

## Operation 1: Filter

**Filter** narrows the data to what matters. It's the difference between "How bad is measles in Africa?" (too broad) and "What happened in East African countries after 2001?" (answerable).

In the blog, I argue that vague questions become specific when you can translate them into operations. Filter is always the first one.

```{r}
#| label: filter-east-africa
#| echo: true

east_africa_post <- measles |>
  filter(region == "East Africa", year >= 2002)

east_africa_post
```

Now we're looking at 5 countries over 22 years — a focused question with a manageable answer.

You can also filter for extremes to find the interesting cases:

```{r}
#| label: filter-high-burden
#| echo: true

measles |>
  filter(cases_per_million > 2000) |>
  count(country, sort = TRUE)
```

DRC, Niger, and Nigeria dominate the high-burden observations. That's not surprising if you know the epidemiology — but it's the kind of pattern that filter surfaces immediately.

## Operation 2: Group

**Grouping** turns a flat table into structured comparisons. "Cases per million across all countries" is a list. "Cases per million *by region and era*" is a comparison — and comparisons are where stories live.

```{r}
#| label: group-by-region-era
#| echo: true

by_region_era <- measles |>
  group_by(region, era) |>
  summarize(
    mean_cases = round(mean(cases_per_million)),
    median_cases = round(median(cases_per_million)),
    n_observations = n(),
    .groups = "drop"
  )

by_region_era
```

Every region declined after 2001. But the Sahel and Central Africa declined less than East and West Africa. That variation is a finding — and it's the entry point for the diagnostic question: *why did some regions improve faster than others?*

## Operation 3: Summarize

**Summarize** reduces data to the numbers that matter — means, medians, rates, percentages. In consulting, this is the step that turns a spreadsheet into a briefing.

The most useful summary here: before-and-after comparison by country.

```{r}
#| label: country-summary
#| echo: true

country_summary <- measles |>
  group_by(country, era) |>
  summarize(mean_cases = round(mean(cases_per_million)), .groups = "drop") |>
  pivot_wider(names_from = era, values_from = mean_cases) |>
  rename(
    pre  = `Pre-Initiative (1990-2001)`,
    post = `Post-Initiative (2002-2023)`
  ) |>
  mutate(pct_change = round((post - pre) / pre * 100, 1)) |>
  arrange(pct_change)

country_summary
```

Every country shows decline. But the range is wide — Kenya improved by roughly 87%, while Nigeria improved by about 55%. That 30+ percentage point gap *is* the story. A single average across all countries would hide it completely.

This is why the blog emphasizes *specificity*. "Measles declined in Africa" is true but useless. "Measles declined everywhere, but 3x faster in East Africa than in the Sahel" is specific, and it raises the obvious next question.

## Operation 4: Compare

**Compare** is where findings become arguments. The first three operations are descriptive — they show what's in the data. Comparison adds the analytical layer: *relative to what?*

```{r}
#| label: compare-income
#| echo: true

income_comparison <- measles |>
  group_by(income_group, era) |>
  summarize(mean_cases = round(mean(cases_per_million)), .groups = "drop") |>
  pivot_wider(names_from = era, values_from = mean_cases) |>
  rename(
    pre  = `Pre-Initiative (1990-2001)`,
    post = `Post-Initiative (2002-2023)`
  ) |>
  mutate(pct_change = round((post - pre) / pre * 100, 1))

income_comparison
```

Lower-middle-income countries showed a sharper decline than low-income countries. That's not definitive — income is correlated with dozens of other factors — but it's a structured comparison that opens the door to a causal argument.

We can also classify countries by their decline speed and compare the two groups:

```{r}
#| label: compare-decline-speed
#| echo: true

country_summary |>
  mutate(decline_speed = if_else(pct_change < -70, "Fast decline", "Slow decline")) |>
  group_by(decline_speed) |>
  summarize(
    countries = paste(country, collapse = ", "),
    avg_pct_change = round(mean(pct_change), 1)
  )
```

The fast decliners average something like -80%, the slow decliners around -55%. That gap is the gap between a vaccination campaign success story and a persistent burden — and it's the kind of comparison that belongs in a policy brief.

## Visualizing the Findings

Once you've filtered, grouped, summarized, and compared, the visualization decisions are already made. You know what you want to show. The chart is just the delivery vehicle.

### The Heatmap

This is the chart type that inspired the [africa-measles project](https://bk-advisors.github.io/africa-measles/). A heatmap shows *everything* — every country, every year — and lets the reader's eye find the pattern.

```{r}
#| label: heatmap
#| echo: true
#| fig-width: 10
#| fig-height: 6

ggplot(measles, aes(
  x = year,
  y = fct_reorder(country, cases_per_million, .fun = mean, .desc = TRUE),
  fill = cases_per_million
)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_gradient(low = "#3E9B6E", high = "#E24A3F",
                      name = "Cases per\nmillion") +
  geom_vline(xintercept = 2001, linetype = "dashed", color = "white", linewidth = 0.8) +
  annotate("text", x = 2001, y = 0.5, label = "2001 Measles Initiative",
           hjust = 1.05, size = 3, color = "grey30") +
  labs(
    title = "Measles Cases per Million Across African Countries, 1990-2023",
    subtitle = "The 2001 WHO/UNICEF Measles Initiative marks a visible inflection point",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), panel.grid = element_blank())
```

The vertical dashed line at 2001 is the diagnosis made visible. Before it: a wall of red. After it: a gradient toward green — but uneven. That unevenness is the story.

### Before vs. After

The grouped bar chart makes the comparison explicit: every country improved, but the magnitude varies.

```{r}
#| label: bar-chart
#| echo: true
#| fig-width: 10
#| fig-height: 5

country_summary |>
  pivot_longer(cols = c(pre, post), names_to = "period", values_to = "mean_cases") |>
  mutate(period = if_else(period == "pre", "Pre-Initiative", "Post-Initiative"),
         period = fct_relevel(period, "Pre-Initiative")) |>
  ggplot(aes(x = fct_reorder(country, mean_cases, .desc = TRUE),
             y = mean_cases, fill = period)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("Pre-Initiative" = "#E24A3F", "Post-Initiative" = "#3E9B6E")) +
  labs(
    title = "Average Measles Cases: Before vs. After the 2001 Initiative",
    subtitle = "Every country improved — but the magnitude varies widely",
    x = NULL, y = "Cases per million", fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

### Decline Ranking

The dot plot ranks countries by their improvement — making it immediately clear who benefited most and who lagged.

```{r}
#| label: dot-plot
#| echo: true
#| fig-width: 9
#| fig-height: 5

ggplot(country_summary, aes(x = pct_change,
                            y = fct_reorder(country, pct_change))) +
  geom_segment(aes(xend = 0, yend = country), color = "grey70") +
  geom_point(size = 4, color = "#005CB9") +
  geom_text(aes(label = paste0(pct_change, "%")), hjust = -0.3, size = 3.5) +
  labs(
    title = "Percent Change in Measles Cases After 2001 Initiative",
    subtitle = "Ranking reveals which countries benefited most — and which lagged",
    x = "% change in mean cases per million", y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## The Diagnostic Layer

Everything above is **descriptive** — what happened. This section adds the **diagnostic** layer: *why* did it happen?

We can compare specific snapshot years to see how the decline played out over time by country.

```{r}
#| label: diagnostic
#| echo: true

diagnostic <- measles |>
  filter(year %in% c(2000, 2001, 2010, 2020)) |>
  select(country, region, income_group, year, cases_per_million) |>
  pivot_wider(names_from = year, values_from = cases_per_million, names_prefix = "y") |>
  mutate(
    change_2001_to_2010 = round((y2010 - y2001) / y2001 * 100, 1),
    change_2001_to_2020 = round((y2020 - y2001) / y2001 * 100, 1)
  )

diagnostic |> arrange(change_2001_to_2020)
```

The fastest improvers cluster in East Africa — countries with relatively stronger health systems and sustained campaign coverage. The slowest improvers tend to be conflict-affected (DRC) or in the Sahel (Niger, Mali), where campaign delivery is harder.

That's not proof of causation. It's a *causal hypothesis grounded in data*. And that's exactly where the diagnostic quadrant lives: not at certainty, but at structured reasoning about mechanisms.

## The Predictive Layer

The final mode: **what happens next?** If the diagnostic question asks "why did this happen?", the predictive question asks "what will happen if current trends continue — or if they don't?"

```{r}
#| label: projections
#| echo: true
#| warning: false

projections <- measles |>
  filter(year >= 2010) |>
  group_by(country) |>
  do({
    model <- lm(log(cases_per_million) ~ year, data = .)
    future_years <- tibble(year = 2024:2030)
    future_years$projected_cases <- round(exp(predict(model, future_years)))
    future_years$country <- unique(.$country)
    future_years
  }) |>
  ungroup()

projections |>
  filter(year == 2030) |>
  arrange(projected_cases) |>
  select(country, projected_cases)
```

Simple exponential extrapolation — not a sophisticated epidemiological model, but enough to illustrate the principle. Now, what happens if progress slows?

```{r}
#| label: scenario-comparison
#| echo: true
#| fig-width: 10
#| fig-height: 7

scenario_slowdown <- projections |>
  mutate(scenario = "Current trend") |>
  bind_rows(
    projections |>
      mutate(projected_cases = round(projected_cases * 1.5),
             scenario = "50% slower decline")
  )

ggplot(scenario_slowdown, aes(x = year, y = projected_cases, color = scenario)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~country, scales = "free_y") +
  scale_color_manual(values = c("Current trend" = "#3E9B6E", "50% slower decline" = "#E24A3F")) +
  labs(
    title = "Projected Measles Cases per Million, 2024-2030",
    subtitle = "What happens if progress slows? The gap widens fast in high-burden countries.",
    x = NULL, y = "Projected cases per million", color = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold"),
    legend.position = "top"
  )
```

In high-burden countries like Nigeria and DRC, the gap between scenarios is large — meaning sustained campaign effort matters *more*, not less, as case counts fall. This is the kind of finding that turns a recommendation into a funded program: not "keep vaccinating" but "if coverage drops, here's what happens."

## How the Operations Map to the Four Modes

| Operation | What it does | Storytelling mode |
|-----------|-------------|-------------------|
| **Filter** | Narrow to what matters | Descriptive — defines scope |
| **Group** | Create structure for comparison | Descriptive — adds dimension |
| **Summarize** | Reduce to key numbers | Descriptive — distills the finding |
| **Compare** | Establish relative differences | Diagnostic — asks *why* and *relative to what* |
| **Project** | Extrapolate or model forward | Predictive — asks *what if* |

The prescriptive mode isn't a data operation — it's the *process* that ties the others together. It's the five-step workflow from the blog: find your question, make it measurable, analyze, synthesize, choose your form.

You don't need advanced statistics to tell a complete data story. You need these four operations, a clear question, and the willingness to ask *why* and *what comes next*.

---

*Companion to [My Approach to Storytelling with Data](blog.html). Full R script: [`tutorial-analysis.R`](tutorial-analysis.R).*
