---
title: "Analysis Techniques for Storytelling with Data"
subtitle: "A hands-on R tutorial: filter, group, summarize, compare — and the questions they answer"
author: "Matthew Kuch"
date: 2026-03-28
format:
  html:
    theme: [cosmo, blog-styles.scss]
    toc: true
    toc-depth: 2
    code-copy: true
    code-fold: false
---

This tutorial walks through the core analysis techniques from the [companion blog post](blog.html) — using R and the tidyverse to demonstrate how four simple data operations (filter, group, summarize, compare) map to four modes of data storytelling (descriptive, diagnostic, prescriptive, predictive).

We'll work with real WHO measles surveillance data: 46 African countries from 1980 to 2023, sourced from the WHO Global Health Observatory (cases) and World Bank (population). If you want to run the code yourself, grab the standalone script: [`tutorial-analysis.R`](tutorial-analysis.R).

## Setup

```{r}
#| label: setup
#| echo: true
#| message: false

library(tidyverse)
library(scales)
```

## Loading the Data

This is real data — not a simulation. The WHO reports measles cases by country-year; we've combined that with World Bank population estimates to get a standardized rate: cases per million people.

```{r}
#| label: load-data
#| echo: true

measles_raw <- read_csv("data/africa_measles.csv", show_col_types = FALSE)

glimpse(measles_raw)
```

`r nrow(measles_raw)` rows: 46 countries across up to 44 years. Each row has the country, its ISO code, the year, raw case counts, population, and cases per million.

### Enriching the Data

The raw data has country and year but no region or income classification. Adding these lets us group and compare — which is where the interesting stories live.

```{r}
#| label: enrich-data
#| echo: true

measles <- measles_raw |>
  mutate(
    region = case_when(
      country %in% c("Algeria") ~ "North Africa",
      country %in% c("Benin", "Burkina Faso", "Cabo Verde", "Cote d'Ivoire",
                      "Gambia", "Ghana", "Guinea", "Guinea-Bissau", "Liberia",
                      "Mali", "Mauritania", "Niger", "Nigeria", "Senegal",
                      "Sierra Leone", "Togo") ~ "West Africa",
      country %in% c("Cameroon", "Central African Republic", "Chad", "Congo",
                      "Democratic Republic of the Congo", "Equatorial Guinea",
                      "Gabon", "Sao Tome and Principe") ~ "Central Africa",
      country %in% c("Burundi", "Comoros", "Eritrea", "Ethiopia", "Kenya",
                      "Madagascar", "Malawi", "Mauritius", "Mozambique",
                      "Rwanda", "Seychelles", "Uganda",
                      "United Republic of Tanzania", "Zambia",
                      "Zimbabwe") ~ "East Africa",
      country %in% c("Angola", "Botswana", "Eswatini", "Lesotho",
                      "Namibia", "South Africa") ~ "Southern Africa"
    ),
    income_group = case_when(
      country %in% c("Algeria", "Botswana", "Equatorial Guinea", "Gabon",
                      "Mauritius", "Namibia", "South Africa") ~ "Upper-middle",
      country %in% c("Angola", "Cabo Verde", "Cameroon", "Comoros", "Congo",
                      "Cote d'Ivoire", "Eswatini", "Ghana", "Kenya",
                      "Lesotho", "Mauritania", "Nigeria",
                      "Sao Tome and Principe", "Senegal",
                      "United Republic of Tanzania", "Zambia",
                      "Zimbabwe") ~ "Lower-middle",
      TRUE ~ "Low"
    ),
    era = if_else(year < 2001, "Pre-Initiative (1980-2000)",
                  "Post-Initiative (2001-2023)")
  )
```

Now each row carries its UN sub-region, World Bank income group, and a before/after label anchored at 2001 — when the WHO/UNICEF Measles & Rubella Initiative launched coordinated vaccination campaigns across Africa.

## Operation 1: Filter

**Filter** narrows the data to what matters. It's the difference between "How bad is measles in Africa?" (too broad) and "What happened in East African countries after 2001?" (answerable).

In the blog, I argue that vague questions become specific when you can translate them into operations. Filter is always the first one.

```{r}
#| label: filter-east-africa
#| echo: true

east_africa_post <- measles |>
  filter(region == "East Africa", year >= 2001)

east_africa_post |>
  select(country, year, cases_per_million) |>
  head(10)
```

Now we're looking at 15 countries over 23 years — a focused question with a manageable answer.

You can also filter for extremes to find the interesting cases:

```{r}
#| label: filter-high-burden
#| echo: true

measles |>
  filter(cases_per_million > 2000) |>
  count(country, sort = TRUE)
```

The high-burden observations cluster in specific countries — and almost exclusively in the pre-2001 era. That's not surprising if you know the epidemiology, but it's the kind of pattern that filter surfaces immediately.

## Operation 2: Group

**Grouping** turns a flat table into structured comparisons. "Cases per million across all countries" is a list. "Cases per million *by region and era*" is a comparison — and comparisons are where stories live.

```{r}
#| label: group-by-region-era
#| echo: true

by_region_era <- measles |>
  group_by(region, era) |>
  summarize(
    mean_cases = round(mean(cases_per_million, na.rm = TRUE)),
    median_cases = round(median(cases_per_million, na.rm = TRUE)),
    n_observations = n(),
    .groups = "drop"
  )

by_region_era
```

Every region declined after 2001. But the magnitude varies — some regions dropped by over 80%, others by far less. That variation is a finding, and it's the entry point for the diagnostic question: *why did some regions improve faster than others?*

## Operation 3: Summarize

**Summarize** reduces data to the numbers that matter — means, medians, rates, percentages. In consulting, this is the step that turns a spreadsheet into a briefing.

The most useful summary here: before-and-after comparison by country.

```{r}
#| label: country-summary
#| echo: true

country_summary <- measles |>
  group_by(country, era) |>
  summarize(mean_cases = round(mean(cases_per_million, na.rm = TRUE)),
            .groups = "drop") |>
  pivot_wider(names_from = era, values_from = mean_cases) |>
  rename(
    pre  = `Pre-Initiative (1980-2000)`,
    post = `Post-Initiative (2001-2023)`
  ) |>
  filter(!is.na(pre), !is.na(post)) |>
  mutate(pct_change = round((post - pre) / pre * 100, 1)) |>
  arrange(pct_change)

country_summary
```

The range is wide. Some countries saw 90%+ declines; others improved far less. That gap *is* the story. A single average across all countries would hide it completely.

This is why the blog emphasizes *specificity*. "Measles declined in Africa" is true but useless. "Measles declined everywhere, but dramatically faster in some regions than others" is specific, and it raises the obvious next question.

## Operation 4: Compare

**Compare** is where findings become arguments. The first three operations are descriptive — they show what's in the data. Comparison adds the analytical layer: *relative to what?*

```{r}
#| label: compare-income
#| echo: true

income_comparison <- measles |>
  group_by(income_group, era) |>
  summarize(mean_cases = round(mean(cases_per_million, na.rm = TRUE)),
            .groups = "drop") |>
  pivot_wider(names_from = era, values_from = mean_cases) |>
  rename(
    pre  = `Pre-Initiative (1980-2000)`,
    post = `Post-Initiative (2001-2023)`
  ) |>
  mutate(pct_change = round((post - pre) / pre * 100, 1))

income_comparison
```

That's not definitive — income is correlated with dozens of other factors — but it's a structured comparison that opens the door to a causal argument.

We can also classify countries by their decline speed and compare the two groups:

```{r}
#| label: compare-decline-speed
#| echo: true

country_summary |>
  mutate(decline_speed = case_when(
    pct_change < -80 ~ "Fast decline (>80%)",
    pct_change < -50 ~ "Moderate decline (50-80%)",
    TRUE ~ "Slow decline (<50%)"
  )) |>
  group_by(decline_speed) |>
  summarize(
    n_countries = n(),
    countries = paste(country, collapse = ", "),
    avg_pct_change = round(mean(pct_change), 1)
  )
```

The gap between fast and slow decliners is the gap between a vaccination campaign success story and a persistent burden — and it's the kind of comparison that belongs in a policy brief.

## Visualizing the Findings

Once you've filtered, grouped, summarized, and compared, the visualization decisions are already made. You know what you want to show. The chart is just the delivery vehicle.

### The Heatmap

This is the chart type that inspired the [africa-measles project](https://bk-advisors.github.io/africa-measles/). A heatmap shows *everything* — every country, every year — and lets the reader's eye find the pattern.

The code below is adapted from the production heatmap published on LinkedIn. It uses the BK Advisors brand palette, orders countries by pre-intervention disease burden, and caps the color scale at 5,000 cases per million (with `scales::squish` clamping outliers to the maximum color rather than dropping them).

```{r}
#| label: bka-theme-setup
#| echo: true

# ---- BKA brand colors (inlined for portability) ----
bka_navy      <- "#242852"
bka_gray      <- "#606060"
bka_blue      <- "#005CB9"
bka_light_blue <- "#ACCBF9"
bka_green     <- "#83BD00"
bka_teal      <- "#3E9B6E"
bka_amber     <- "#F8A623"
bka_orange    <- "#FA7650"
bka_red       <- "#E24A3F"

# ---- Multi-stop gradient: cool (safe) → warm (danger) ----
measles_gradient <- c(
  colorRampPalette(c(bka_light_blue, bka_blue, bka_green, bka_amber))(15),
  colorRampPalette(c(bka_amber, bka_orange, bka_red))(5)
)
```

```{r}
#| label: heatmap
#| echo: true
#| fig-width: 10
#| fig-height: 12

# ---- Sort countries by average pre-2001 case rate (highest at top) ----
country_order <- measles |>
  filter(year < 2001) |>
  group_by(country) |>
  summarise(avg_rate = mean(cases_per_million, na.rm = TRUE), .groups = "drop") |>
  arrange(avg_rate) |>
  pull(country)

measles_ordered <- measles |>
  mutate(country = factor(country, levels = country_order))

# ---- Build the heatmap ----
ggplot(measles_ordered, aes(x = year, y = country, fill = cases_per_million)) +
  geom_tile(colour = "white", linewidth = 0.4,
            width = 0.9, height = 0.9) +
  scale_fill_gradientn(
    colours = measles_gradient,
    limits = c(0, 5000),
    breaks = seq(0, 5000, by = 1000),
    labels = c("0", "1K", "2K", "3K", "4K", "5K"),
    na.value = "#F5F5F5",
    name = "Reported cases per million population",
    oob = scales::squish,
    guide = guide_colourbar(
      barwidth = 15, barheight = 0.6,
      title.position = "top",
      title.hjust = 0.5,
      ticks = TRUE, nbin = 50
    )
  ) +
  scale_x_continuous(expand = c(0, 0),
                     breaks = seq(1980, 2020, by = 5)) +
  geom_vline(xintercept = 2001, color = bka_navy,
             linewidth = 1.2, alpha = 0.8) +
  annotate("text", label = "WHO/UNICEF\nMeasles Initiative",
           x = 2001, y = length(country_order) + 0.5,
           hjust = -0.05, vjust = 1,
           size = 3.5, fontface = "bold",
           color = bka_navy) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Mass immunization campaigns cut African measles\ncases by over 90% — but outbreaks persist",
    subtitle = "Reported measles cases per million population by country, 1980-2023",
    x = NULL, y = NULL,
    caption = paste0(
      "Source: WHO Global Health Observatory (cases) & World Bank (population) | ",
      "Countries with 15+ years of reporting shown\n",
      "Visualization: BK Advisors"
    )
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_blank(),
    axis.text.y = element_text(size = 8, face = "bold", hjust = 1),
    axis.text.x = element_text(size = 8),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.title = element_text(size = 9, face = "bold", color = bka_navy),
    legend.text = element_text(size = 8, color = bka_gray),
    plot.title = element_text(face = "bold", color = bka_navy, lineheight = 1.15),
    plot.subtitle = element_text(color = bka_gray, margin = margin(t = 2, b = 14)),
    plot.caption = element_text(color = bka_gray, size = 7, hjust = 1),
    plot.title.position = "plot",
    plot.caption.position = "plot",
    plot.margin = margin(15, 15, 10, 5)
  )
```

The vertical line at 2001 is the diagnosis made visible. Before it: a wall of orange and red. After it: a gradient toward blue — but uneven. That unevenness is the story. The gray tiles are missing data — countries that didn't report in certain years — which is itself informative about surveillance infrastructure.

### Before vs. After

The grouped bar chart makes the comparison explicit. We'll show the 15 highest-burden countries (by pre-initiative average) to keep it readable.

```{r}
#| label: bar-chart
#| echo: true
#| fig-width: 10
#| fig-height: 6

top_15 <- country_summary |>
  arrange(desc(pre)) |>
  head(15)

top_15 |>
  pivot_longer(cols = c(pre, post), names_to = "period", values_to = "mean_cases") |>
  mutate(period = if_else(period == "pre", "Pre-Initiative", "Post-Initiative"),
         period = fct_relevel(period, "Pre-Initiative")) |>
  ggplot(aes(x = fct_reorder(country, mean_cases, .desc = TRUE),
             y = mean_cases, fill = period)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("Pre-Initiative" = bka_red,
                                "Post-Initiative" = bka_teal)) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Average Measles Cases: Before vs. After the 2001 Initiative",
    subtitle = "Top 15 highest-burden countries — every one improved, but the magnitude varies widely",
    x = NULL, y = "Cases per million", fill = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", color = bka_navy),
    plot.subtitle = element_text(color = bka_gray),
    plot.title.position = "plot",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8)
  )
```

### Decline Ranking

The dot plot ranks countries by their improvement — making it immediately clear who benefited most and who lagged. Again, top 20 for readability.

```{r}
#| label: dot-plot
#| echo: true
#| fig-width: 9
#| fig-height: 7

top_20_change <- country_summary |>
  head(20)

ggplot(top_20_change, aes(x = pct_change,
                           y = fct_reorder(country, pct_change))) +
  geom_segment(aes(xend = 0, yend = country), color = "grey70") +
  geom_point(size = 4, color = bka_blue) +
  geom_text(aes(label = paste0(pct_change, "%")), hjust = -0.3, size = 3.2) +
  labs(
    title = "Percent Change in Measles Cases After 2001 Initiative",
    subtitle = "Top 20 fastest-improving countries",
    x = "% change in mean cases per million", y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", color = bka_navy),
    plot.subtitle = element_text(color = bka_gray),
    plot.title.position = "plot"
  )
```

## The Diagnostic Layer

Everything above is **descriptive** — what happened. This section adds the **diagnostic** layer: *why* did it happen?

We can compare specific snapshot years to see how the decline played out over time by region.

```{r}
#| label: diagnostic
#| echo: true

diagnostic <- measles |>
  filter(year %in% c(2000, 2005, 2010, 2020)) |>
  group_by(region, year) |>
  summarize(
    mean_rate = round(mean(cases_per_million, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = year, values_from = mean_rate, names_prefix = "y")

diagnostic
```

The fastest improvers tend to be in regions with relatively stronger health systems and sustained campaign coverage. The slowest improvers are often in conflict-affected areas or where campaign delivery is logistically harder.

That's not proof of causation. It's a *causal hypothesis grounded in data*. And that's exactly where the diagnostic quadrant lives: not at certainty, but at structured reasoning about mechanisms.

## The Predictive Layer

The final mode: **what happens next?** If the diagnostic question asks "why did this happen?", the predictive question asks "what will happen if current trends continue — or if they don't?"

```{r}
#| label: projections
#| echo: true
#| warning: false

# Pick 10 high-burden countries for projection
projection_countries <- country_summary |>
  arrange(desc(pre)) |>
  head(10) |>
  pull(country)

projections <- measles |>
  filter(year >= 2010, country %in% projection_countries) |>
  group_by(country) |>
  do({
    model <- lm(log(pmax(cases_per_million, 1)) ~ year, data = .)
    future_years <- tibble(year = 2024:2030)
    future_years$projected_cases <- round(exp(predict(model, future_years)))
    future_years$country <- unique(.$country)
    future_years
  }) |>
  ungroup()

projections |>
  filter(year == 2030) |>
  arrange(projected_cases) |>
  select(country, projected_cases)
```

Simple exponential extrapolation — not a sophisticated epidemiological model, but enough to illustrate the principle. Now, what happens if progress slows?

```{r}
#| label: scenario-comparison
#| echo: true
#| fig-width: 10
#| fig-height: 7

scenario_slowdown <- projections |>
  mutate(scenario = "Current trend") |>
  bind_rows(
    projections |>
      mutate(projected_cases = round(projected_cases * 1.5),
             scenario = "50% slower decline")
  )

ggplot(scenario_slowdown, aes(x = year, y = projected_cases, color = scenario)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~country, scales = "free_y", ncol = 5) +
  scale_color_manual(values = c("Current trend" = bka_teal,
                                 "50% slower decline" = bka_red)) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Projected Measles Cases per Million, 2024-2030",
    subtitle = "What happens if progress slows? The gap widens fast in high-burden countries.",
    x = NULL, y = "Projected cases per million", color = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(face = "bold", color = bka_navy),
    plot.subtitle = element_text(color = bka_gray),
    plot.title.position = "plot",
    strip.text = element_text(face = "bold", size = 8),
    legend.position = "top"
  )
```

In high-burden countries, the gap between scenarios is large — meaning sustained campaign effort matters *more*, not less, as case counts fall. This is the kind of finding that turns a recommendation into a funded program: not "keep vaccinating" but "if coverage drops, here's what happens."

## How the Operations Map to the Four Modes

| Operation | What it does | Storytelling mode |
|-----------|-------------|-------------------|
| **Filter** | Narrow to what matters | Descriptive — defines scope |
| **Group** | Create structure for comparison | Descriptive — adds dimension |
| **Summarize** | Reduce to key numbers | Descriptive — distills the finding |
| **Compare** | Establish relative differences | Diagnostic — asks *why* and *relative to what* |
| **Project** | Extrapolate or model forward | Predictive — asks *what if* |

The prescriptive mode isn't a data operation — it's the *process* that ties the others together. It's the five-step workflow from the blog: find your question, make it measurable, analyze, synthesize, choose your form.

You don't need advanced statistics to tell a complete data story. You need these four operations, a clear question, and the willingness to ask *why* and *what comes next*.

---

*Companion to [My Approach to Storytelling with Data](blog.html). Full R script: [`tutorial-analysis.R`](tutorial-analysis.R).*
