---
title: "Chapter 1: Introduction to Causal Analysis"
subtitle: "Situational assessment, managerial intervention, and why correlation ≠ causation"
---

## Learning Objectives

::: {.learning-objectives}
After this chapter, you will be able to:

- Distinguish between **situational assessment** and **managerial intervention**
- Explain why **correlation does not imply causation** with real MNH examples
- Identify three classic causal thinking traps: confounding, reverse causation, and selection bias
- Describe the basic elements of a **causal model** (nodes, arrows, directed graphs)
- Articulate why causal thinking is essential for large-scale pooled investment decisions in maternal and newborn health programs
:::

## Key Concepts

::: {.key-concept}
#### Situational Assessment
Using evidence to reason about the current state of a system — determining which variables are in which states, and why. This is the "What is happening?" question every manager faces.
:::

::: {.key-concept .green}
#### Managerial Intervention
A decision about which objects to control or change — the choice of *what* to act on, before deciding *how* to act. Resources are always limited, so every intervention involves trade-offs.
:::

::: {.key-concept .red}
#### The Three Causal Thinking Traps

| Trap | What Happens | MNH Example |
|---|---|---|
| **Confounding** | Third variable drives both X and Y | Disease burden drives both worker deployment and mortality |
| **Reverse Causation** | Y causes X, not X causes Y | Mortality attracts funding, not vice versa |
| **Selection Bias** | Non-random sample distorts relationship | Only high-capacity hospitals report data |
:::

::: {.key-concept .teal}
#### Directed Acyclic Graph (DAG)
A causal diagram with no cycles — you cannot follow arrows from any node back to itself. An arrow from A to B means that changing A can change B, but changing B does *not* change A. This is the mathematical foundation of causal modeling.
:::

## Lecture Slides

::: {.slide-embed}
<iframe src="../ch01-intro/index.html"></iframe>
<div class="slide-embed-footer">
<a href="../ch01-intro/index.html" target="_blank">Open slides full-screen &rarr;</a>
</div>
:::

## R Code Workshop

This chapter introduces the core R packages for causal analysis: `dagitty` for DAG specification and analysis, `ggdag` for visualization, and `ggplot2` for general plotting.

### Setting Up the Environment

```{r}
#| label: setup
#| echo: true

# Install packages (run once)
# install.packages(c("dagitty", "ggdag", "ggplot2", "bnlearn"))

# Load packages
library(dagitty)    # DAG specification and analysis
library(ggdag)      # DAG visualization with ggplot2
library(ggplot2)    # General plotting
```

### A Simple MNH Causal Model

```{r}
#| label: simple-dag
#| echo: true
#| fig-width: 8
#| fig-height: 4

# Define a simple 3-node DAG
mnh_simple <- dagitty('dag {
  Investment [pos="0,0"]
  Coverage [pos="1,0"]
  NMR [pos="2,0"]

  Investment -> Coverage
  Coverage -> NMR
}')

# Visualize
ggdag(mnh_simple) +
  theme_dag() +
  labs(title = "Simple MNH Causal Model",
       subtitle = "Investment → Coverage → Neonatal Mortality Rate")
```

### A Confounded Model

```{r}
#| label: confounded-dag
#| echo: true
#| fig-width: 9
#| fig-height: 5

# Hospital delivery and mortality: confounded by complication severity
confounded_dag <- dagitty('dag {
  Complication_Severity [pos="1,0"]
  Hospital_Delivery [pos="0,1"]
  Neonatal_Mortality [pos="2,1"]

  Complication_Severity -> Hospital_Delivery
  Complication_Severity -> Neonatal_Mortality
  Hospital_Delivery -> Neonatal_Mortality
}')

# Visualize with exposure and outcome
ggdag_status(confounded_dag,
             exposure = "Hospital_Delivery",
             outcome = "Neonatal_Mortality") +
  theme_dag() +
  scale_color_manual(values = c("exposure" = "#83BD00",
                                "outcome" = "#E24A3F",
                                "other" = "#F8A623")) +
  labs(title = "Confounding in Hospital Delivery Data")
```

### Finding Adjustment Sets

```{r}
#| label: adjustment-sets
#| echo: true

# What must we control for to estimate the causal effect
# of hospital delivery on mortality?
adjustmentSets(confounded_dag,
               exposure = "Hospital_Delivery",
               outcome = "Neonatal_Mortality")
# Returns: { Complication_Severity }
```

To estimate the true causal effect of hospital delivery on neonatal mortality, we must **control for complication severity**. Without this adjustment, the correlation between hospital delivery and mortality is biased by confounding.

### A Richer MNH Model

```{r}
#| label: rich-dag
#| echo: true
#| fig-width: 10
#| fig-height: 6

# MNH initiative theory of change
mnh_toc <- dagitty('dag {
  Budget [pos="0,1"]
  Workforce [pos="1,0"]
  Equipment [pos="1,1"]
  Data_Systems [pos="1,2"]
  Quality [pos="2,1"]
  NMR [pos="3,1"]
  Baseline_Burden [pos="1.5,2.5"]

  Budget -> Workforce
  Budget -> Equipment
  Budget -> Data_Systems
  Workforce -> Quality
  Equipment -> Quality
  Data_Systems -> Quality
  Quality -> NMR
  Baseline_Burden -> NMR
  Baseline_Burden -> Budget
}')

# Identify the adjustment set for Budget → NMR
adjustmentSets(mnh_toc,
               exposure = "Budget",
               outcome = "NMR")

# Visualize
ggdag(mnh_toc) +
  theme_dag() +
  labs(title = "MNH Initiative: Theory of Change as a DAG")
```

## Key Takeaways

::: {.key-takeaway}
**Causal modeling serves two purposes:** situational assessment (What is happening?) and managerial intervention (What should I do?).

**Correlation ≠ causation.** Confounding, reverse causation, and selection bias all create misleading associations.

**A causal model** makes your assumptions explicit, testable, and actionable — essential for allocating large-scale program funding effectively.
:::

## Looking Ahead

Next in **Chapter 2 — Qualitative Causal Models**, we learn the formal language of causal diagrams:

- Four types of nodes (probabilistic, objective, strategic option, function)
- Directed links with [+] and [−] signs
- The three fundamental triplet structures: serial, diverging, and converging
- How to determine whether two variables are causally independent (d-separation)
