---
title: "Chapter 2: Qualitative Causal Models"
subtitle: "Nodes, directed links, triplet structures, and d-separation"
---

## Learning Objectives

::: {.learning-objectives}
After this chapter, you will be able to:

- Identify the **four types of nodes** in a causal model and when each is used
- Assign **signed directed links** ([+]/[-]) to represent qualitative causal relationships
- Construct **conditional certainty tables** for non-monotonic relationships
- Recognize the three fundamental **triplet structures**: serial (chain), diverging (fork), converging (collider)
- Apply **d-separation** criteria to determine when information flows between variables
- Build a complete **qualitative DAG** for an MNH program's investment model
:::

## Key Concepts

::: {.key-concept}
#### The Four Node Types

A qualitative causal model uses four types of nodes: **Probabilistic nodes** (ovals) represent variables with uncertain states determined by nature. **Objective nodes** (hexagons) represent goals or performance metrics the decision-maker wants to optimize -- they are always leaf nodes. **Strategic option nodes** (rectangles) represent deliberate choices made by a decision-maker -- they are root nodes. **Function nodes** (chevrons) represent variables whose state is deterministically determined by their parents with no residual uncertainty.
:::

::: {.key-concept .green}
#### Signed Directed Links

An arrow from node A to node B means that A is a direct cause of B -- changing A can change B while holding all other parents of B fixed. The sign **[+]** indicates that increasing A increases B, while **[-]** indicates that increasing A decreases B. The overall sign of a directed path is the product of the individual link signs, where [+] x [+] = [+], [-] x [-] = [+], and [+] x [-] = [-].
:::

::: {.key-concept .orange}
#### The Three Triplet Structures

The three fundamental triplet structures govern all information flow in a DAG. In a **serial triplet** (chain, A -> B -> C), B is a mediator and conditioning on B blocks the path. In a **diverging triplet** (fork, A <- B -> C), B is a common cause and conditioning on B removes the spurious correlation. In a **converging triplet** (collider, A -> B <- C), the path is naturally blocked but conditioning on B *opens* it, creating collider bias through the "explaining away" effect.
:::

::: {.key-concept .red}
#### d-Separation

Two nodes X and Y are **d-separated** given a conditioning set Z if every path between them is blocked. A path is blocked if it contains a chain or fork whose middle node is in Z, or a collider whose middle node (and all descendants) is not in Z. d-Separation implies conditional independence -- these are testable predictions of the model. If data violates an implied independence, the model must be revised.
:::

## Lecture Slides

::: {.slide-embed}
<iframe src="../ch02-qualitative-models/index.html"></iframe>
<div class="slide-embed-footer">
<a href="../ch02-qualitative-models/index.html" target="_blank">Open slides full-screen &rarr;</a>
</div>
:::

## R Code Workshop

This chapter builds and analyzes the full MNH qualitative DAG in R, using `dagitty` for DAG specification and d-separation testing, and `ggdag` for visualization.

### Define the Full MNH DAG

We specify the comprehensive 10-node qualitative DAG with positional coordinates for layout, representing the MNH program's theory of change from budget allocation through mediating variables to mortality outcomes.

```{r}
#| label: mnh-dag-define
#| echo: true

library(dagitty)
library(ggdag)
library(ggplot2)

# Define the comprehensive MNH qualitative DAG
mnh_dag <- dagitty('dag {
  Budget          [pos="0,1"]
  Workforce       [pos="1.5,0"]
  Training        [pos="1.5,1"]
  Equipment       [pos="1.5,2"]
  Referral_System [pos="1.5,3"]
  Coverage        [pos="3,3"]
  Quality         [pos="3,1"]
  Baseline_Burden [pos="4,3"]
  NMR             [pos="4.5,0.5"]
  MMR             [pos="4.5,1.5"]

  Budget -> Workforce
  Budget -> Training
  Budget -> Equipment
  Budget -> Referral_System
  Training -> Quality
  Workforce -> Quality
  Equipment -> Quality
  Referral_System -> Coverage
  Coverage -> Quality
  Quality -> NMR
  Quality -> MMR
  Baseline_Burden -> NMR
  Baseline_Burden -> MMR
}')

# Verify it is a valid DAG (acyclic)
isAcyclic(mnh_dag)
```

### Visualize the DAG Color-Coded by Role

We merge node role metadata into the tidy DAG data to enable color-coding by function: decision nodes, mediators, contextual variables, and outcome objectives.

```{r}
#| label: mnh-dag-visualize
#| echo: true
#| fig-width: 11
#| fig-height: 6

# Create a tidy dag object with exposure and outcome
tidy_dag <- tidy_dagitty(mnh_dag)

# Define node roles for coloring
node_roles <- data.frame(
  name = c("Budget", "Workforce", "Training", "Equipment",
           "Referral_System", "Coverage", "Quality",
           "Baseline_Burden", "NMR", "MMR"),
  role = c("Decision", "Mediator", "Mediator", "Mediator",
           "Mediator", "Mediator", "Mediator",
           "Context", "Outcome", "Outcome")
)

# Merge roles into the dag data
dag_data <- left_join(as.data.frame(tidy_dag), node_roles,
                      by = "name")

ggdag(mnh_dag) +
  theme_dag() +
  labs(title = "MNH Program: Qualitative Causal Model",
       subtitle = "10-node DAG for MNH investment decisions") +
  theme(plot.title = element_text(color = "#005CB9", size = 16),
        plot.subtitle = element_text(color = "#666", size = 12))
```

### Test d-Separation

The `dseparated()` function lets us verify the conditional independence predictions implied by our DAG. Each test checks whether information flows between two variables given a conditioning set.

```{r}
#| label: mnh-dsep-tests
#| echo: true

# Test 1: Is Training d-separated from NMR given Quality?
dseparated(mnh_dag, "Training", "NMR", "Quality")
# Expected: TRUE (chain is blocked)

# Test 2: Is Workforce d-separated from Equipment? (unconditional)
dseparated(mnh_dag, "Workforce", "Equipment", c())
# Expected: FALSE (common parent Budget creates fork)

# Test 3: Is Workforce d-separated from Equipment given Budget?
dseparated(mnh_dag, "Workforce", "Equipment", "Budget")
# Expected: TRUE (conditioning on common cause blocks fork)

# Test 4: Is Workforce d-separated from Equipment given Quality?
dseparated(mnh_dag, "Workforce", "Equipment", "Quality")
# Expected: FALSE (conditioning on collider opens path!)

# Test 5: Is Training d-separated from Baseline_Burden?
dseparated(mnh_dag, "Training", "Baseline_Burden", c())
# Expected: TRUE (no connecting path)
```

### Enumerate All Paths

The `paths()` function lists every path between two nodes in the DAG, showing both directed (causal) and non-directed (non-causal) paths. This is useful for understanding the full set of channels through which one variable can influence another.

```{r}
#| label: mnh-paths
#| echo: true

# Find all paths from Budget to NMR
budget_nmr_paths <- paths(mnh_dag,
                          from = "Budget",
                          to = "NMR")
print(budget_nmr_paths)

# Find all paths from Budget to MMR
budget_mmr_paths <- paths(mnh_dag,
                          from = "Budget",
                          to = "MMR")
print(budget_mmr_paths)

# Find paths between Workforce and Equipment (through collider)
workforce_equip_paths <- paths(mnh_dag,
                               from = "Workforce",
                               to = "Equipment")
print(workforce_equip_paths)
```

### Implied Conditional Independencies

The `impliedConditionalIndependencies()` function extracts every testable prediction from the DAG. These are the conditional independencies that must hold in the data if the model is correctly specified.

```{r}
#| label: mnh-implied-ci
#| echo: true

# List all conditional independencies implied by the DAG
implied_cis <- impliedConditionalIndependencies(mnh_dag)
print(implied_cis)

# This tells us every testable prediction of our model.
# For example, if our model is correct, we should observe:
#   - Training _||_ Equipment | Budget
#   - Workforce _||_ Referral_System | Budget
#   - Training _||_ Baseline_Burden
#   - NMR _||_ MMR | Quality, Baseline_Burden
#
# If DHIS2 data violates any of these, we need to revise the DAG.

# Count total number of implied independencies
cat("Total implied conditional independencies:", length(implied_cis), "\n")
```

## Key Takeaways

::: {.key-takeaway}
**Qualitative causal models are the foundation of rigorous analysis.** A DAG encodes your causal beliefs explicitly using four node types and signed directed links.

The three triplet structures -- **chain**, **fork**, and **collider** -- govern all information flow. Chains and forks are blocked by conditioning; colliders are *opened* by conditioning.

**d-Separation** translates graph structure into testable conditional independence predictions.

For MNH programs: the qualitative DAG reveals that People, Products, and Systems investments are **complements** flowing through Quality of Care -- and that conditioning on Quality creates collider bias between investment pillars.
:::

## Looking Ahead

Next in **Chapter 3 -- The MNH Diagnostic Case Study**, we put qualitative models to work in a structured case study:

- Building a DAG from stakeholder interviews through an iterative process
- Translating qualitative beliefs into a causal diagram round by round
- Identifying key assumptions, disagreements, and transmission failures
- Preparing the model for quantification in later chapters
