---
title: "Chapter 4: Quantitative Causal Models"
subtitle: "Conditional probability tables, Bayes' rule, and Bayesian networks"
---

## Learning Objectives

::: {.learning-objectives}
After this chapter, you will be able to:

- Define **probability distributions** over discrete variable states and apply the basic axioms
- Compute **conditional probabilities** and construct **Conditional Probability Tables (CPTs)**
- Apply **Bayes' rule** to update beliefs when new evidence arrives
- Explain the **Causal Markov Condition** and how a DAG factorizes a joint distribution
- Build a **quantitative Bayesian network** in R using `bnlearn` and query it for MNH policy decisions
:::

## Key Concepts

::: {.key-concept}
#### Conditional Probability Tables (CPTs)

A CPT lists P(Child | Parents) for every combination of parent and child states. Each row in the table defines a conditional distribution, and every row must sum to 1. CPTs are the quantitative building blocks of a Bayesian network -- they encode the strength of causal relationships that qualitative models could only describe with [+] and [-] signs. In practice, CPTs are estimated from facility-level data, expert elicitation, or both.
:::

::: {.key-concept .green}
#### Bayes' Rule

Bayes' rule lets us "flip" the conditioning direction: given P(Evidence | Cause), it computes P(Cause | Evidence). The formula is P(A | B) = P(B | A) x P(A) / P(B). The prior P(A) represents what we believed before seeing evidence, the likelihood P(B | A) represents how probable the evidence is if the hypothesis is true, and the posterior P(A | B) represents our updated belief after incorporating evidence. This is the fundamental operation of situational assessment -- reasoning backward from what we observe to what is likely true.
:::

::: {.key-concept .orange}
#### The Causal Markov Condition

Each variable in a DAG is conditionally independent of its non-descendants given its parents. This means the full joint distribution factorizes as a product of CPTs: P(X1, X2, ..., Xn) = Product of P(Xi | Parents(Xi)). This factorization reduces computational complexity from exponential to manageable -- instead of specifying millions of probabilities, the causal structure lets us specify just a few hundred.
:::

::: {.key-concept .teal}
#### Bayesian Network

A Bayesian network equals a DAG plus a CPT for every node. It allows us to compute exact probabilities for any query, update beliefs with Bayes' rule when evidence arrives, and compare interventions using the do-operator. The do-operator severs incoming arrows to the intervention variable and sets it to a fixed value, giving us the causal effect rather than a mere correlation.
:::

## Lecture Slides

::: {.slide-embed}
<iframe src="../ch04-quantitative-models/index.html"></iframe>
<div class="slide-embed-footer">
<a href="../ch04-quantitative-models/index.html" target="_blank">Open slides full-screen &rarr;</a>
</div>
:::

## R Code Workshop

This chapter builds a complete Bayesian network in R using `bnlearn`, specifies CPTs for each node, and demonstrates forward queries, interventional queries, and sensitivity analysis.

### Define the Network Structure

We use `bnlearn`'s model string syntax to specify the DAG. The notation `[Node|Parent1:Parent2]` means a node is conditional on its parents, while `[Node]` alone denotes a root node with no parents.

```{r}
#| label: bn-structure
#| echo: true

library(bnlearn)

# Define the DAG structure using bnlearn's model string syntax
# [Node] means no parents; [Node|Parent1:Parent2] means conditional on parents
dag <- model2network(
  "[Staffing][Equipment][Quality|Staffing:Equipment][NMR|Quality]"
)

# Inspect the structure
dag
nodes(dag)    # "Staffing" "Equipment" "Quality" "NMR"
arcs(dag)     # Shows the 3 directed edges
```

### Specify CPTs

Each CPT is defined as an array with named dimensions. Root nodes get simple marginal distributions, while child nodes get conditional distributions indexed by all parent state combinations.

```{r}
#| label: bn-cpts
#| echo: true

# Define CPTs using arrays (states must match exactly)
cpt_staffing <- array(c(0.40, 0.60), dim = 2,
  dimnames = list(Staffing = c("Adequate", "Low")))

cpt_equipment <- array(c(0.35, 0.65), dim = 2,
  dimnames = list(Equipment = c("Available", "Unavailable")))

cpt_quality <- array(
  c(0.85, 0.15,  # Adequate + Available
    0.50, 0.50,  # Low + Available
    0.55, 0.45,  # Adequate + Unavailable
    0.20, 0.80), # Low + Unavailable
  dim = c(2, 2, 2),
  dimnames = list(
    Quality   = c("Good", "Poor"),
    Staffing  = c("Adequate", "Low"),
    Equipment = c("Available", "Unavailable")))

cpt_nmr <- array(
  c(0.85, 0.15,  # Good quality
    0.40, 0.60), # Poor quality
  dim = c(2, 2),
  dimnames = list(NMR = c("Low", "High"), Quality = c("Good", "Poor")))

# Combine into a fitted BN
bn_mnh <- custom.fit(dag, list(
  Staffing = cpt_staffing, Equipment = cpt_equipment,
  Quality = cpt_quality, NMR = cpt_nmr))
```

### Query the Network

The `cpquery()` function uses likelihood weighting to estimate conditional probabilities from the network. For exact inference, we convert the network to a junction tree using the `gRain` package.

```{r}
#| label: bn-query
#| echo: true

# Query: P(NMR = High | Staffing = Low)
cpquery(bn_mnh,
  event  = (NMR == "High"),
  evidence = (Staffing == "Low"),
  n = 1e6)
# Expected: ~0.44

# Query: P(NMR = High | Staffing = Adequate, Equipment = Available)
cpquery(bn_mnh,
  event  = (NMR == "High"),
  evidence = (Staffing == "Adequate" & Equipment == "Available"),
  n = 1e6)
# Expected: ~0.22

# Exact inference using grain (junction tree)
library(gRain)
bn_grain <- as.grain(bn_mnh)
querygrain(bn_grain, nodes = "NMR")
# Returns marginal P(NMR) across all facilities
```

### Visualize the Network

Two visualization options are available: `bnlearn`'s built-in plot using Rgraphviz, and `ggdag` for publication-quality figures built on ggplot2.

```{r}
#| label: bn-visualize
#| echo: true
#| fig-width: 10
#| fig-height: 5

# Option 1: bnlearn's built-in plot (uses Rgraphviz)
graphviz.plot(bn_mnh, main = "MNH Bayesian Network",
  shape = "rectangle",
  highlight = list(nodes = "NMR", col = "#E24A3F", fill = "#E24A3F"))

# Option 2: ggdag for publication-quality figures
library(ggdag)
library(dagitty)

mnh_dag <- dagitty('dag {
  Staffing [pos="0,0"]
  Equipment [pos="2,0"]
  Quality [pos="1,1"]
  NMR [pos="1,2"]
  Staffing -> Quality
  Equipment -> Quality
  Quality -> NMR
}')

ggdag(mnh_dag) +
  theme_dag() +
  labs(title = "MNH Causal Model (4 Nodes)")
```

### Interventional Queries

The `mutilated()` function implements the do-operator -- it severs incoming arrows to the intervention variable and sets it to a fixed value. This lets us compare the causal effect of different interventions side by side.

```{r}
#| label: bn-intervention
#| echo: true

# Simulate do(Staffing = Adequate) by mutilating the network
bn_do_adequate <- mutilated(bn_mnh, evidence = list(Staffing = "Adequate"))

# P(NMR | do(Staffing = Adequate))
querygrain(as.grain(bn_do_adequate), nodes = "NMR")

# Compare scenarios
scenarios <- data.frame(
  Scenario = c("Status Quo", "do(Staff=Adequate)",
               "do(Equip=Available)", "do(Both)"),
  P_NMR_High = c(
    querygrain(as.grain(bn_mnh),
               nodes = "NMR")$NMR["High"],
    querygrain(as.grain(
      mutilated(bn_mnh, list(Staffing = "Adequate"))),
               nodes = "NMR")$NMR["High"],
    querygrain(as.grain(
      mutilated(bn_mnh, list(Equipment = "Available"))),
               nodes = "NMR")$NMR["High"],
    querygrain(as.grain(
      mutilated(bn_mnh, list(Staffing = "Adequate",
                              Equipment = "Available"))),
               nodes = "NMR")$NMR["High"]))

# Display comparison
scenarios$Reduction <- scenarios$P_NMR_High[1] - scenarios$P_NMR_High
print(scenarios)
```

## Key Takeaways

::: {.key-takeaway}
**Quantitative causal models = DAG + CPTs = Bayesian Network.** They allow us to compute exact probabilities, update beliefs with Bayes' rule, and compare interventions using the do-operator.

**Probability fundamentals:** States must be MECE; probabilities must sum to 1. CPTs encode the quantitative relationships in the DAG.

**Bayes' rule** lets us reason backward from evidence to causes -- essential for situational assessment.

**The Causal Markov Condition** simplifies computation from exponential to manageable. **Sensitivity analysis** reveals which interventions have the largest expected impact -- in the MNH model, staffing has a larger marginal effect than equipment, but the two are synergistic.
:::

## Looking Ahead

Next in **Chapter 5 -- Situational Analysis**, we turn to the first major application of quantitative causal models:

- Updating beliefs with multiple evidence streams -- combining facility data, surveys, and expert judgment
- Diagnostic inference -- given observed outcomes, which upstream causes are most likely?
- The explaining away phenomenon -- how one piece of evidence changes the relevance of another
- A complete district diagnosis using Bayesian inference on the MNH causal model
