---
title: "The Case for HPV Vaccination in Papua New Guinea"
subtitle: "Cervical cancer is devastating women across Papua New Guinea. A proven vaccine could save nearly 100,000 lives by 2070."
author: "Matthew Kuch"
date: today
format: closeread-html
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(scales)
library(showtext)

# Load data
source("R/data.R")

# Add system fonts
font_add_google("Source Sans Pro", "source")
showtext_auto()

# Define color palette (matching NYT-inspired design)
colors <- list(
  primary = "#121f2b",
  danger = "#c41d3a",
  danger_light = "#fdf6f6",
  success = "#1a7f4b",
  success_light = "#f4faf7",
  text = "#1a1a1a",
  text_secondary = "#555",
  text_light = "#777",
  grid = "#f0f0f0",
  bar_default = "#c4c9cf"
)

# Custom theme inspired by Cairo, Tufte, and NYT
theme_story <- function() {

  theme_minimal(base_family = "source", base_size = 12) +
    theme(
      plot.title = element_text(size = 14, face = "bold", color = colors$text,
                                margin = margin(b = 5)),
      plot.subtitle = element_text(size = 10, color = colors$text_light,
                                   margin = margin(b = 15)),
      plot.caption = element_text(size = 8, color = colors$text_light,
                                  hjust = 0, margin = margin(t = 10)),
      axis.title = element_text(size = 9, color = colors$text_light),
      axis.text = element_text(size = 9, color = colors$text_light),
      axis.line = element_blank(),
      panel.grid.major.y = element_line(color = colors$grid, linewidth = 0.5),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      legend.position = "none",
      plot.margin = margin(20, 20, 20, 20),
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA)
    )
}
```

::: {.cr-section}

## A Silent Crisis {focus-on="global-stat"}

In 2022, cervical cancer claimed the lives of **348,709 women** worldwide. More than half of these deaths occurred in the Asia-Pacific region.

:::{#global-stat .sticky}
```{r}
#| label: global-deaths
#| fig-width: 6
#| fig-height: 4

# Create a simple stat display
ggplot() +

  annotate("text", x = 0.5, y = 0.6, label = format(global_deaths$total, big.mark = ","),
           size = 20, fontface = "bold", color = colors$danger, family = "source") +
  annotate("text", x = 0.5, y = 0.35,
           label = "women died from cervical cancer\nglobally in 2022",
           size = 5, color = colors$text_secondary, family = "source",
           lineheight = 1.2) +
  xlim(0, 1) + ylim(0, 1) +
  theme_void() +
  theme(plot.background = element_rect(fill = "white", color = NA))
```
:::

## The Asia-Pacific Burden {focus-on="regional-stat"}

The Asia-Pacific region bears a disproportionate burden, accounting for **197,217 deaths** - 56.6% of the global total.

:::{#regional-stat .sticky}
```{r}
#| label: regional-breakdown
#| fig-width: 7
#| fig-height: 4

regional_df <- tibble(
  region = c("Asia-Pacific", "Other Regions"),
  deaths = c(global_deaths$asia_pacific, global_deaths$other_regions),
  percent = c(global_deaths$asia_pacific_percent, global_deaths$other_regions_percent)
)

ggplot(regional_df, aes(x = reorder(region, -deaths), y = deaths,
                         fill = region == "Asia-Pacific")) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(format(deaths, big.mark = ","), "\n(", percent, "%)")),
            vjust = -0.3, size = 4, family = "source", color = colors$text_secondary) +
  scale_fill_manual(values = c("TRUE" = colors$danger, "FALSE" = colors$bar_default)) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Cervical Cancer Deaths by Region (2022)",
    x = NULL,
    y = "Number of deaths",
    caption = "Source: Global Cancer Observatory (GCO, 2022)"
  ) +
  theme_story()
```
:::

## Papua New Guinea's Challenge {focus-on="bar-chart"}

Papua New Guinea has the **second-highest** cervical cancer mortality rate in the Western Pacific region, at **19.9 deaths per 100,000 women**.

:::{#bar-chart .sticky}
```{r}
#| label: mortality-rates
#| fig-width: 8
#| fig-height: 6

ggplot(mortality_rates, aes(x = reorder(country, rate), y = rate,
                             fill = highlight)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = rate), hjust = -0.2, size = 3.5,
            family = "source", color = colors$text_light) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = colors$danger, "FALSE" = colors$bar_default)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Cervical Cancer Mortality in Asia-Pacific (2022)",
    subtitle = "Age-standardized mortality rate per 100,000 women",
    x = NULL,
    y = "Deaths per 100,000",
    caption = "Source: Ueda (2024), Journal of Obstetrics and Gynaecology Research"
  ) +
  theme_story() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(color = colors$grid, linewidth = 0.5)
  )
```
:::

## A Tale of Two Neighbors {focus-on="comparison-chart"}

The disparity becomes stark when comparing PNG to Australia. PNG's incidence rate is **5.5 times higher**, and its mortality rate is nearly **12 times higher**.

:::{#comparison-chart .sticky}
```{r}
#| label: comparison
#| fig-width: 7
#| fig-height: 5

comparison_long <- comparison_data |>
  tidyr::pivot_longer(cols = c(incidence, mortality),
                      names_to = "metric", values_to = "rate") |>
  mutate(metric = factor(metric, levels = c("incidence", "mortality"),
                         labels = c("Incidence (new cases)", "Mortality (deaths)")))

ggplot(comparison_long, aes(x = country, y = rate, fill = interaction(country, metric))) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = rate), position = position_dodge(width = 0.8),
            vjust = -0.3, size = 3.5, family = "source") +
  scale_fill_manual(values = c(
    "Papua New Guinea.Incidence (new cases)" = "#e8a5a5",
    "Papua New Guinea.Mortality (deaths)" = colors$danger,
    "Australia.Incidence (new cases)" = "#a8c8e8",
    "Australia.Mortality (deaths)" = "#4a7eb8"
  )) +
  facet_wrap(~metric, scales = "free_y") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "PNG vs Australia: Cervical Cancer Rates",
    subtitle = "Per 100,000 women, 2022",
    x = NULL,
    y = "Rate per 100,000",
    caption = "Source: Globocan 2022"
  ) +
  theme_story() +
  theme(strip.text = element_text(size = 10, face = "bold"))
```
:::

## The Stakes {focus-on="stakes-stat"}

Without significant intervention, an estimated **60,232 women** in PNG could die from cervical cancer by 2070. Each death represents a mother, daughter, sister - a life cut short.

:::{#stakes-stat .sticky}
```{r}
#| label: stakes
#| fig-width: 6
#| fig-height: 4

ggplot() +
  annotate("text", x = 0.5, y = 0.6,
           label = format(deaths_without_action$by_2070, big.mark = ","),
           size = 20, fontface = "bold", color = colors$danger, family = "source") +
  annotate("text", x = 0.5, y = 0.35,
           label = "women in PNG could die from\ncervical cancer by 2070\nwithout intervention",
           size = 5, color = colors$text_secondary, family = "source",
           lineheight = 1.2) +
  xlim(0, 1) + ylim(0, 1) +
  theme_void() +
  theme(plot.background = element_rect(fill = "white", color = NA))
```
:::

## A Proven Solution {focus-on="solution-stat"}

The HPV vaccine is among the most effective cancer prevention tools available. Gavi estimates that vaccination prevents **17.4 deaths** for every 1,000 girls vaccinated.

:::{#solution-stat .sticky}
```{r}
#| label: solution
#| fig-width: 6
#| fig-height: 5

ggplot() +
  annotate("text", x = 0.5, y = 0.65, label = "17.4",
           size = 25, fontface = "bold", color = colors$success, family = "source") +
  annotate("text", x = 0.5, y = 0.45,
           label = "deaths averted for every\n1,000 girls vaccinated",
           size = 5, color = colors$text_secondary, family = "source",
           lineheight = 1.2) +
  annotate("rect", xmin = 0.1, xmax = 0.9, ymin = 0.05, ymax = 0.3,
           fill = colors$success_light, color = NA) +
  annotate("text", x = 0.5, y = 0.175,
           label = "Single-dose schedule now\nrecommended by WHO",
           size = 4, color = colors$success, family = "source",
           lineheight = 1.3) +
  xlim(0, 1) + ylim(0, 1) +
  theme_void() +
  theme(plot.background = element_rect(fill = "white", color = NA))
```
:::

## Lives That Could Be Saved {focus-on="projection-chart"}

If PNG introduces comprehensive HPV vaccination, projections show that nearly **97,000 lives** could be saved by 2070.

:::{#projection-chart .sticky}
```{r}
#| label: lives-saved
#| fig-width: 8
#| fig-height: 5

ggplot(projected_lives_saved, aes(x = year, y = cumulative_lives_saved)) +
  geom_area(fill = colors$success, alpha = 0.15) +
  geom_line(color = colors$success, linewidth = 1.5) +
  geom_point(data = projected_lives_saved |> filter(year == 2070),
             color = colors$success, size = 4) +
  geom_text(data = projected_lives_saved |> filter(year == 2070),
            aes(label = paste0(format(cumulative_lives_saved, big.mark = ","), " lives saved")),
            hjust = -0.1, vjust = 0.5, size = 4, family = "source",
            fontface = "bold", color = colors$success) +
  scale_x_continuous(breaks = c(2026, 2030, 2040, 2050, 2060, 2070),
                     limits = c(2026, 2080)) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Projected Lives Saved Through HPV Vaccination",
    subtitle = "Papua New Guinea, 2026-2070",
    x = "Year",
    y = "Cumulative lives saved",
    caption = "Based on Gavi estimate of 17.4 deaths averted per 1,000 vaccinated.\nAssumes 80% initial coverage expanding to 90%."
  ) +
  theme_story()
```
:::

## A Sound Investment {focus-on="roi-chart"}

HPV vaccination isn't just good health policy - it's smart economics. UNFPA modeling shows **$19 return** for every $1 invested over 30 years, growing to **$61** over 50 years.

:::{#roi-chart .sticky}
```{r}
#| label: roi
#| fig-width: 7
#| fig-height: 5

ggplot(roi_data, aes(x = period, y = value, fill = !is_investment)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = label), vjust = -0.3, size = 5, family = "source",
            fontface = "bold", color = ifelse(roi_data$is_investment,
                                               colors$text_light, colors$success)) +
  scale_fill_manual(values = c("TRUE" = colors$success, "FALSE" = colors$bar_default)) +
  scale_x_discrete(limits = roi_data$period) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Return on Investment",
    subtitle = "For every $1 invested in HPV vaccination",
    x = NULL,
    y = "Dollars",
    caption = "Source: UNFPA PNG cervical cancer elimination modeling"
  ) +
  theme_story()
```
:::

## PNG Is Ready {focus-on="ready-stat"}

PNG has already piloted HPV vaccination in the National Capital District and Western Highlands, demonstrating its capacity to deliver adolescent vaccination programs.

:::{#ready-stat .sticky}
```{r}
#| label: ready
#| fig-width: 6
#| fig-height: 5

ready_stats <- tibble(
  stat = c("5.5M+", "~97,000"),
  label = c("girls could be vaccinated by 2070", "lives saved")
)

ggplot() +
  annotate("text", x = 0.5, y = 0.85, label = "Building on Experience",
           size = 6, fontface = "bold", color = colors$primary, family = "source") +
  annotate("rect", xmin = 0.15, xmax = 0.45, ymin = 0.6, ymax = 0.72,
           fill = "#f0f7fc", color = NA) +
  annotate("text", x = 0.3, y = 0.66, label = "National Capital District",
           size = 3.5, color = colors$primary, family = "source") +
  annotate("rect", xmin = 0.55, xmax = 0.85, ymin = 0.6, ymax = 0.72,
           fill = "#f0f7fc", color = NA) +
  annotate("text", x = 0.7, y = 0.66, label = "Western Highlands",
           size = 3.5, color = colors$primary, family = "source") +
  annotate("rect", xmin = 0.2, xmax = 0.45, ymin = 0.25, ymax = 0.45,
           fill = colors$success_light, color = NA) +
  annotate("text", x = 0.325, y = 0.38, label = "5.5M+",
           size = 8, fontface = "bold", color = colors$success, family = "source") +
  annotate("text", x = 0.325, y = 0.29, label = "girls vaccinated\nby 2070",
           size = 3, color = colors$text_light, family = "source", lineheight = 1.1) +
  annotate("rect", xmin = 0.55, xmax = 0.8, ymin = 0.25, ymax = 0.45,
           fill = colors$success_light, color = NA) +
  annotate("text", x = 0.675, y = 0.38, label = "~97,000",
           size = 8, fontface = "bold", color = colors$success, family = "source") +
  annotate("text", x = 0.675, y = 0.29, label = "lives saved",
           size = 3, color = colors$text_light, family = "source") +
  xlim(0, 1) + ylim(0, 1) +
  theme_void() +
  theme(plot.background = element_rect(fill = "white", color = NA))
```
:::

## The Time Is Now {focus-on="cta"}

In March 2024, PNG's Senior Executive Management officially recommended HPV vaccine introduction. PNG has signed the WHO's Western Pacific 2023-2030 framework for cervical cancer elimination.

:::{#cta .sticky}
```{r}
#| label: cta
#| fig-width: 6
#| fig-height: 6

checkmarks <- tibble(
  y = c(0.85, 0.72, 0.59, 0.46),
  label = c("SEM recommendation (March 2024)",
            "WHO Western Pacific Framework signatory",
            "Gavi support eligible",
            "Pilot program experience")
)

ggplot() +
  geom_point(data = checkmarks, aes(x = 0.12, y = y),
             shape = 21, fill = colors$success, color = "white", size = 5) +
  geom_text(data = checkmarks, aes(x = 0.15, y = y, label = label),
            hjust = 0, size = 3.5, family = "source", color = colors$text_secondary) +
  annotate("rect", xmin = 0.05, xmax = 0.95, ymin = 0.1, ymax = 0.32,
           fill = colors$primary, color = NA) +
  annotate("text", x = 0.5, y = 0.25,
           label = "Introducing HPV vaccination is not just\na technical immunization decision.",
           size = 3.5, color = "white", alpha = 0.85, family = "source",
           lineheight = 1.3) +
  annotate("text", x = 0.5, y = 0.15,
           label = "It is a gender, equity, and social justice decision.",
           size = 4, fontface = "bold", color = "white", family = "source") +
  xlim(0, 1) + ylim(0, 1) +
  theme_void() +
  theme(plot.background = element_rect(fill = "white", color = NA))
```
:::

:::

---

## Sources

1. Global Cancer Observatory (GCO, 2022)
2. Ueda (2024), Journal of Obstetrics and Gynaecology Research
3. Globocan PNG Factsheet, 2022
4. UNFPA PNG Cervical Cancer Elimination Brief
5. Gavi HPV Impact Report, 2022
6. WHO Western Pacific Region Framework 2023-2030

---

*Built with R and Quarto Closeread*
