---
title: "Chart Reengineering 02 — Vaccination Heatmap (Original Data)"
subtitle: "Faithful recreation of the R Graph Gallery's measles heatmap with BKA styling"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.width = 12, fig.height = 10)
```

## About This Lesson

This is the companion to [01-vaccination-heatmap.qmd](01-vaccination-heatmap.qmd), which adapted the chart to Ethiopian MNH data. Here we recreate the **original chart** using the actual US measles dataset — 51 states across 84 years (1928-2011). Compare the two side by side to see how the same structural pattern works at very different data scales.

Original: [r-graph-gallery.com/vaccination-heatmap.html](https://r-graph-gallery.com/vaccination-heatmap.html)

## The Answer

The original vaccination heatmap, rebuilt with BKA typography and styling:

```{r the-answer}
library(ggplot2)
library(dplyr)
library(readr)
source("../../synthesized-lessons/_common/theme_bka.R")

# Load the original measles dataset from R Graph Gallery
measles <- read_csv("https://raw.githubusercontent.com/holtzy/R-graph-gallery/refs/heads/master/measles_data_1928-2011.csv")

# See Lesson 1.2 (Aesthetics) — factor() with rev() for reverse-alphabetical y-axis
measles <- measles %>%
  mutate(State = factor(State, levels = rev(sort(unique(State)))))

# See Lesson 2.4 (Palettes) — colorRampPalette() creates a smooth multi-stop gradient
# Original chart uses asymmetric split: 20 steps for low-mid, 80 for high
# This compresses the upper range into intense reds — visual urgency for outbreaks
cols <- c(
  colorRampPalette(c("#e7f0fa", "#c9e2f6", "#95cbee",
                     "#0099dc", "#4ab04a", "#ffd73e",
                     "#eec73a", "#e29421"))(20),
  colorRampPalette(c("#e29421", "#f05336", "#ce472e"))(80)
)

ggplot(measles, aes(x = Year, y = State, fill = value)) +
  # See Lesson 1.4 (Geometries) — geom_tile() with white borders and sub-1.0 dimensions
  geom_tile(colour = "white", linewidth = 0.5,
            width = 0.9, height = 0.9) +    # Tufte: white space between tiles improves readability
  # See Lesson 2.4 (Palettes) — scale_fill_gradientn() for multi-stop continuous gradients
  scale_fill_gradientn(
    colours = cols,
    limits = c(0, 4000),
    breaks = seq(0, 4000, by = 1000),
    labels = c("0k", "1k", "2k", "3k", "4k"),
    na.value = rgb(246/255, 246/255, 246/255),
    name = "Cases per million",
    # See Lesson 2.5 (Legend Styling) — guide_colourbar() for continuous fill legends
    guide = guide_colourbar(
      barwidth = 15, barheight = 0.6,
      title.position = "top",
      title.hjust = 0.5,
      ticks = TRUE, nbin = 50
    )
  ) +
  # See Lesson 1.7 (Scales) — expand = c(0,0) removes padding; breaks set tick marks
  scale_x_continuous(expand = c(0, 0),
                     breaks = seq(1930, 2010, by = 10),
                     limits = c(1928, 2012)) +
  # See Lesson 3.2 (Annotations) — geom_vline() marks a policy milestone
  geom_vline(xintercept = 1963, color = bka_colors$title_dark,
             linewidth = 0.5) +    # Cairo: annotations are essential, not optional
  # See Lesson 3.2 (Annotations) — annotate("text") for floating labels
  annotate("text", label = "Vaccine\nintroduced",
           x = 1963, y = 53,
           hjust = -0.1, vjust = 0.9,
           size = 3, family = "Lato",
           color = bka_colors$title_dark) +
  # See Lesson 3.1 (Titles & Captions) — Minto-style title states the insight
  labs(
    title = "Measles vaccination in 1963 nearly eliminated outbreaks within a decade",
    subtitle = "Measles cases per million by US state, 1928-2011",
    x = NULL, y = NULL,
    caption = "Source: Project Tycho / R Graph Gallery | Recreated with BKA styling"
  ) +
  # See Lesson 2.2 (Custom Themes) — theme_bka() + heatmap-specific overrides
  theme_bka() +
  theme(
    panel.grid = element_blank(),    # Tufte: no gridlines for heatmaps — tiles ARE the data
    axis.line = element_blank(),
    axis.text.y = element_text(size = 6, family = "Lato", hjust = 1),    # Few: small text for 51 states
    axis.text.x = element_text(size = 8),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.title = element_text(size = 9, face = "bold",
                                color = bka_colors$title_dark),
    legend.text = element_text(size = 8, color = bka_colors$subtitle_gray),
    plot.margin = margin(15, 15, 25, 15)
  )
```

## Key Insight

> **The most powerful heatmaps encode a clear before/after story.** The 1963 vaccine line splits the chart into two unmistakable color regimes — hot reds on the left, cool blues on the right. When your data has a structural break, a single reference line can do more narrative work than paragraphs of text.

## Building Blocks

### Step 1: Load and inspect the original data

The dataset has ~4,300 rows: 51 states across 84 years, with measles cases per million as the value.

```{r step-1-data}
library(ggplot2)
library(dplyr)
library(readr)
source("../../synthesized-lessons/_common/theme_bka.R")

measles <- read_csv("https://raw.githubusercontent.com/holtzy/R-graph-gallery/refs/heads/master/measles_data_1928-2011.csv")

# See Lesson 1.2 (Aesthetics) — factor() controls y-axis display order
measles <- measles %>%
  mutate(State = factor(State, levels = rev(sort(unique(State)))))

glimpse(measles)
```

The `value` column ranges from 0 to over 3,000 cases per million, with many NAs (states that didn't report in early years).

### Step 2: Basic heatmap — 51 states x 84 years

At this scale (4,284 tiles), you immediately see why heatmaps work: no other geometry can show this much data in one view.

```{r step-2-basic}
# See Lesson 1.4 (Geometries) — geom_tile() creates a heatmap grid
ggplot(measles, aes(x = Year, y = State, fill = value)) +
  geom_tile() +
  labs(title = "Step 2: Basic heatmap — default gradient hides the story") +
  theme_bka() +
  theme(axis.text.y = element_text(size = 5))
```

The default gradient compresses everything into a uniform blue wash. We need a gradient that makes outbreaks visually alarming.

### Step 3: Tile spacing at scale

With 4,000+ tiles, the white borders create a subtle texture that helps the eye track rows and columns.

```{r step-3-spacing}
# See Lesson 1.4 (Geometries) — colour, linewidth, width, height for tile appearance
ggplot(measles, aes(x = Year, y = State, fill = value)) +
  geom_tile(colour = "white", linewidth = 0.5,
            width = 0.9, height = 0.9) +    # Tufte: white space between tiles
  labs(title = "Step 3: White borders separate each state-year cell") +
  theme_bka() +
  theme(panel.grid = element_blank(),
        axis.text.y = element_text(size = 5))
```

### Step 4: The original multi-stop gradient

This is the signature design choice. The asymmetric 20/80 split means most of the gradient's visual range is spent on the high values (outbreaks), while the low values (post-vaccine era) compress into cool blues.

```{r step-4-gradient}
# See Lesson 2.4 (Palettes) — colorRampPalette() for smooth multi-stop gradients
# Original chart: 20 steps for 8 anchor colors (low-mid), 80 steps for 3 anchor colors (high)
cols <- c(
  colorRampPalette(c("#e7f0fa", "#c9e2f6", "#95cbee",
                     "#0099dc", "#4ab04a", "#ffd73e",
                     "#eec73a", "#e29421"))(20),
  colorRampPalette(c("#e29421", "#f05336", "#ce472e"))(80)
)

# See Lesson 2.4 (Palettes) — scale_fill_gradientn() maps data to gradient
ggplot(measles, aes(x = Year, y = State, fill = value)) +
  geom_tile(colour = "white", linewidth = 0.5,
            width = 0.9, height = 0.9) +
  scale_fill_gradientn(
    colours = cols,
    limits = c(0, 4000),
    breaks = seq(0, 4000, by = 1000),
    labels = c("0k", "1k", "2k", "3k", "4k"),
    na.value = rgb(246/255, 246/255, 246/255),    # Light gray for missing data
    name = "Cases per million"
  ) +
  labs(title = "Step 4: Asymmetric gradient — outbreaks pop, post-vaccine fades") +
  theme_bka() +
  theme(panel.grid = element_blank(),
        axis.text.y = element_text(size = 5))
```

Why asymmetric? With a symmetric gradient, the rare extreme values (3,000+ per million) would dominate and the moderate outbreaks (500-1,500) would look identical to low values. The 20/80 split ensures the *entire* pre-vaccine era reads as visually intense.

### Step 5: The 1963 reference line

This is the annotation that transforms the chart from "a wall of color" into "a story about vaccination."

```{r step-5-reference-line}
ggplot(measles, aes(x = Year, y = State, fill = value)) +
  geom_tile(colour = "white", linewidth = 0.5,
            width = 0.9, height = 0.9) +
  scale_fill_gradientn(
    colours = cols, limits = c(0, 4000),
    breaks = seq(0, 4000, by = 1000),
    labels = c("0k", "1k", "2k", "3k", "4k"),
    na.value = rgb(246/255, 246/255, 246/255),
    name = "Cases per million"
  ) +
  scale_x_continuous(expand = c(0, 0),
                     breaks = seq(1930, 2010, by = 10),
                     limits = c(1928, 2012)) +
  # See Lesson 3.2 (Annotations) — geom_vline() for reference lines
  geom_vline(xintercept = 1963, color = bka_colors$title_dark,
             linewidth = 0.5) +    # Cairo: annotations are essential, not optional
  # See Lesson 3.2 (Annotations) — annotate("text") for floating labels
  annotate("text", label = "Vaccine\nintroduced",
           x = 1963, y = 53,
           hjust = -0.1, vjust = 0.9,
           size = 3, family = "Lato",
           color = bka_colors$title_dark) +
  labs(title = "Step 5: The 1963 line splits the chart into two color regimes") +
  theme_bka() +
  theme(panel.grid = element_blank(),
        axis.text.y = element_text(size = 5))
```

The label sits at `y = 53`, slightly above the 51 state rows. This is the same `annotate("text")` technique from **Lesson 3.2 — Annotations**.

### Step 6: Final polish with BKA styling

```{r step-6-final}
ggplot(measles, aes(x = Year, y = State, fill = value)) +
  geom_tile(colour = "white", linewidth = 0.5,
            width = 0.9, height = 0.9) +
  scale_fill_gradientn(
    colours = cols,
    limits = c(0, 4000),
    breaks = seq(0, 4000, by = 1000),
    labels = c("0k", "1k", "2k", "3k", "4k"),
    na.value = rgb(246/255, 246/255, 246/255),
    name = "Cases per million",
    # See Lesson 2.5 (Legend Styling) — guide_colourbar() for continuous legends
    guide = guide_colourbar(
      barwidth = 15, barheight = 0.6,
      title.position = "top", title.hjust = 0.5,
      ticks = TRUE, nbin = 50
    )
  ) +
  scale_x_continuous(expand = c(0, 0),
                     breaks = seq(1930, 2010, by = 10),
                     limits = c(1928, 2012)) +
  geom_vline(xintercept = 1963, color = bka_colors$title_dark, linewidth = 0.5) +
  annotate("text", label = "Vaccine\nintroduced",
           x = 1963, y = 53,
           hjust = -0.1, vjust = 0.9,
           size = 3, family = "Lato",
           color = bka_colors$title_dark) +
  # See Lesson 3.1 (Titles & Captions) — Minto-style title states the insight
  labs(
    title = "Measles vaccination in 1963 nearly eliminated outbreaks within a decade",
    subtitle = "Measles cases per million by US state, 1928-2011",
    x = NULL, y = NULL,
    caption = "Source: Project Tycho / R Graph Gallery | Recreated with BKA styling"
  ) +
  # See Lesson 2.2 (Custom Themes) — theme_bka() + heatmap-specific overrides
  theme_bka() +
  theme(
    panel.grid = element_blank(),    # Tufte: no gridlines for heatmaps
    axis.line = element_blank(),
    axis.text.y = element_text(size = 6, family = "Lato", hjust = 1),    # Few: readable at 51 rows
    axis.text.x = element_text(size = 8),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.title = element_text(size = 9, face = "bold",
                                color = bka_colors$title_dark),
    legend.text = element_text(size = 8, color = bka_colors$subtitle_gray),
    plot.margin = margin(15, 15, 25, 15)
  )
```

## Scale Comparison: 8 Regions vs 51 States

The same heatmap pattern works at very different scales:

| Dimension | MNH Version (Lesson 01) | Original (This Lesson) |
|-----------|------------------------|----------------------|
| Y-axis rows | 8 regions | 51 states |
| X-axis columns | 11 years | 84 years |
| Total tiles | 88 | ~4,300 |
| Y-axis text size | 9pt | 6pt |
| Figure dimensions | 10 x 7 | 12 x 10 |
| Gradient range | 150-500 MMR | 0-4,000 cases/million |
| Gradient split | 15/5 (BKA colors) | 20/80 (original colors) |

The key design adaptation for scale: smaller y-axis text, taller figure, and wider x-axis range.

## Quick Reference

| Technique | Function | Key Arguments | Lesson |
|-----------|----------|---------------|--------|
| Load CSV from URL | `readr::read_csv()` | URL string | — |
| Heatmap grid | `geom_tile()` | `colour`, `linewidth`, `width`, `height` | 1.4 — Geometries |
| Map fill to continuous | `aes(fill = value)` | Inside `ggplot()` | 1.2 — Aesthetics |
| Y-axis ordering | `factor(levels = rev(...))` | `rev(sort(unique(...)))` | 1.2 — Aesthetics |
| Multi-stop gradient | `scale_fill_gradientn()` | `colours`, `limits`, `breaks`, `labels` | 2.4 — Palettes |
| Generate gradient | `colorRampPalette()` | Hex vectors, `(n)` steps | 2.4 — Palettes |
| NA tile color | `na.value = rgb(...)` | Inside `scale_fill_gradientn()` | 2.4 — Palettes |
| Reference line | `geom_vline()` | `xintercept`, `color` | 3.2 — Annotations |
| Floating text | `annotate("text")` | `x`, `y`, `label`, `hjust` | 3.2 — Annotations |
| Colorbar legend | `guide_colourbar()` | `barwidth`, `barheight`, `nbin` | 2.5 — Legend Styling |
| Remove axis padding | `scale_x_continuous(expand = c(0,0))` | `limits`, `breaks` | 1.7 — Scales |
| Minto title | `labs(title = "...")` | Insight-first phrasing | 3.1 — Titles & Captions |

## Your Turn

The original chart uses the author's custom color palette. Try replacing it with a BKA-branded gradient using the same anchor colors from Lesson 01. Does the BKA version tell the same story as effectively?

```{r your-turn, eval=FALSE}
# Replace the original gradient with BKA colors
# Hint: use the same anchor points from 01-vaccination-heatmap.qmd
bka_gradient <- c(
  colorRampPalette(c("___", "___", "___", "___"))(20),
  colorRampPalette(c("___", "___", "___"))(80)
)

ggplot(measles, aes(x = Year, y = State, fill = value)) +
  geom_tile(colour = "white", linewidth = 0.5,
            width = 0.9, height = 0.9) +
  scale_fill_gradientn(
    colours = ___,
    limits = c(0, 4000),
    breaks = seq(0, 4000, by = 1000),
    labels = c("0k", "1k", "2k", "3k", "4k"),
    na.value = "#F5F5F5",
    name = "Cases per million",
    guide = guide_colourbar(
      barwidth = 15, barheight = 0.6,
      title.position = "top", title.hjust = 0.5
    )
  ) +
  scale_x_continuous(expand = c(0, 0),
                     breaks = seq(1930, 2010, by = 10),
                     limits = c(1928, 2012)) +
  geom_vline(xintercept = 1963, color = bka_colors$title_dark) +
  annotate("text", label = "Vaccine\nintroduced",
           x = 1963, y = 53,
           hjust = -0.1, vjust = 0.9,
           size = 3, family = "Lato",
           color = bka_colors$title_dark) +
  labs(
    title = "___",
    subtitle = "Measles cases per million by US state, 1928-2011",
    x = NULL, y = NULL,
    caption = "___"
  ) +
  theme_bka() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_blank(),
    axis.text.y = element_text(size = 6, family = "Lato", hjust = 1),
    legend.position = "bottom",
    legend.direction = "horizontal"
  )
```
