---
title: "Chart Reengineering 03 — Measles Vaccination Heatmap for Africa"
subtitle: "Real WHO data showing how mass immunization campaigns transformed a continent"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.width = 12, fig.height = 14)
```

## About This Lesson

Lessons [01](01-vaccination-heatmap.qmd) and [02](02-vaccination-heatmap-original.qmd) used synthetic MNH data and the original US measles dataset. This lesson tells the **real story of measles vaccination in Africa** using WHO surveillance data and World Bank population estimates — 46 countries across 44 years.

The landmark year is **2001**, when the WHO/UNICEF **Measles & Rubella Initiative** launched targeted supplementary immunization activities (SIAs) across Africa. Within a decade, reported measles cases dropped by over 90% in most countries.

### Data Sources

The pre-processed CSV (`data/africa_measles.csv`) was built from two public APIs:

| Source | Indicator | What It Provides |
|--------|-----------|-----------------|
| [WHO Global Health Observatory](https://www.who.int/data/gho) | `WHS3_62` | Measles reported cases by country, 1980-2023 |
| [World Bank Open Data](https://data.worldbank.org) | `SP.POP.TOTL` | Total population by country, 1980-2023 |

To regenerate the data from scratch, run `source("../data/prep_africa_measles.R")` (requires `jsonlite`).

### Techniques You Will Practice

| Technique | Function | From Lesson |
|-----------|----------|-------------|
| Heatmap grid | `geom_tile()` | 1.4 — Geometries |
| Multi-stop color gradient | `scale_fill_gradientn()` | 2.4 — Palettes |
| Squish outliers into gradient range | `oob = scales::squish` | 2.4 — Palettes |
| Y-axis ordering with factors | `factor(levels = ...)` | 1.2 — Aesthetics |
| Vertical reference line | `geom_vline()` | 3.2 — Annotations |
| Floating text label | `annotate("text")` | 3.2 — Annotations |
| Minto-style insight title | `labs()` | 3.1 — Titles & Captions |
| Continuous color legend | `guide_colourbar()` | 2.5 — Legend Styling |
| Axis formatting | `scale_x_continuous()` | 1.7 — Scales |
| Custom theme + overrides | `theme_bka()` | 2.2 — Custom Themes |

## The Answer

A publication-ready heatmap of measles cases per million across 46 African countries from 1980-2023, with a reference line marking the 2001 WHO/UNICEF Measles Initiative:

```{r the-answer}
library(ggplot2)
library(dplyr)
source("../../synthesized-lessons/_common/theme_bka.R")

# ---- Load pre-processed data ----
# Built from WHO GHO API (cases) + World Bank API (population)
# To regenerate: source("../data/prep_africa_measles.R")
africa_measles <- read.csv("../data/africa_measles.csv")

# ---- Sort countries by average pre-2001 case rate (highest at top) ----
# See Lesson 1.2 (Aesthetics) — factor() with custom levels for y-axis ordering
country_order <- africa_measles %>%
  filter(year < 2001) %>%
  group_by(country) %>%
  summarise(avg_rate = mean(cases_per_million, na.rm = TRUE), .groups = "drop") %>%
  arrange(avg_rate) %>%
  pull(country)

africa_measles <- africa_measles %>%
  mutate(country = factor(country, levels = country_order))

# ---- BKA multi-stop gradient ----
# See Lesson 2.4 (Palettes) — colorRampPalette() creates smooth multi-stop gradients
# Asymmetric split: 15 steps for low-mid range, 5 for high = visual urgency at top end
measles_gradient <- c(
  colorRampPalette(c("#ACCBF9", "#005CB9", "#83BD00", "#F8A623"))(15),
  colorRampPalette(c("#F8A623", "#FA7650", "#E24A3F"))(5)
)

# ---- The heatmap ----
ggplot(africa_measles, aes(x = year, y = country, fill = cases_per_million)) +
  # See Lesson 1.4 (Geometries) — geom_tile() with white borders
  geom_tile(colour = "white", linewidth = 0.4,
            width = 0.9, height = 0.9) +    # Tufte: white space between tiles
  # See Lesson 2.4 (Palettes) — scale_fill_gradientn() for multi-stop gradients
  scale_fill_gradientn(
    colours = measles_gradient,
    limits = c(0, 5000),
    breaks = seq(0, 5000, by = 1000),
    labels = c("0", "1K", "2K", "3K", "4K", "5K"),
    na.value = "#F5F5F5",
    name = "Reported cases per million population",
    oob = scales::squish,    # Squish outliers beyond 5000 to the max color
    # See Lesson 2.5 (Legend Styling) — guide_colourbar() for continuous legends
    guide = guide_colourbar(
      barwidth = 15, barheight = 0.6,
      title.position = "top",
      title.hjust = 0.5,
      ticks = TRUE, nbin = 50
    )
  ) +
  # See Lesson 1.7 (Scales) — expand = c(0,0) removes padding
  scale_x_continuous(expand = c(0, 0),
                     breaks = seq(1980, 2020, by = 5)) +
  # See Lesson 3.2 (Annotations) — geom_vline() marks a policy milestone
  geom_vline(xintercept = 2001, color = bka_colors$title_dark,
             linewidth = 1.2, alpha = 0.8) +    # Cairo: annotations are essential, not optional
  # See Lesson 3.2 (Annotations) — annotate("text") for floating labels
  annotate("text", label = "WHO/UNICEF\nMeasles Initiative",
           x = 2001, y = length(country_order) + 0.5,
           hjust = -0.05, vjust = 1,
           size = 4, fontface = "bold", family = "Lato",
           color = bka_colors$title_dark) +
  coord_cartesian(clip = "off") +    # Prevent annotation text from being clipped
  # See Lesson 3.1 (Titles & Captions) — Minto-style title states the insight
  labs(
    title = "Mass immunization campaigns cut African measles cases by over 90%\nbut outbreaks persist where coverage gaps remain",
    subtitle = "Reported measles cases per million population by country, 1980-2023",
    x = NULL, y = NULL,
    caption = paste0(
      "Source: WHO Global Health Observatory (cases) & World Bank (population) | ",
      "Countries with 15+ years of reporting shown\n",
      "Original inspiration: r-graph-gallery.com/vaccination-heatmap.html"
    )
  ) +
  # See Lesson 2.2 (Custom Themes) — theme_bka() + heatmap-specific overrides
  theme_bka() +
  theme(
    panel.grid = element_blank(),    # Tufte: no gridlines for heatmaps
    axis.line = element_blank(),
    axis.text.y = element_text(size = 9, face = "bold", family = "Lato", hjust = 1),    # Few: readable with many rows
    axis.text.x = element_text(size = 8),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.title = element_text(size = 9, face = "bold",
                                color = bka_colors$title_dark),
    legend.text = element_text(size = 8, color = bka_colors$subtitle_gray),
    plot.title = element_text(lineheight = 1.15),
    plot.margin = margin(15, 15, 25, 15)
  )
```

## Key Insight

> **Real-world data tells a more complex story than synthetic data.** The US measles heatmap shows a clean before/after break. Africa's version reveals the same dramatic decline, but also the reality of periodic outbreaks when vaccination coverage slips — a pattern invisible in tidy synthetic datasets. Missing data (gray tiles) is itself informative, showing gaps in surveillance infrastructure.

## Building Blocks

### Step 1: Load and inspect the data

The CSV was pre-built from WHO and World Bank APIs (see `../data/prep_africa_measles.R`). Each row is one country-year with reported cases, population, and the per-million rate.

```{r step-1-load}
library(ggplot2)
library(dplyr)
source("../../synthesized-lessons/_common/theme_bka.R")

africa_measles <- read.csv("../data/africa_measles.csv")

cat("Countries:", length(unique(africa_measles$country)), "\n")
cat("Year range:", min(africa_measles$year), "-", max(africa_measles$year), "\n")
cat("Cases per million range:", round(min(africa_measles$cases_per_million)),
    "-", round(max(africa_measles$cases_per_million)), "\n")
```

46 countries, 1980-2023. The cases_per_million variable ranges from 0 to over 30,000 — a massive range that we'll handle with `oob = scales::squish`.

### Step 2: Order countries by disease burden

Alphabetical order is the default, but it tells no story. Sorting by pre-2001 average case rate puts the highest-burden countries at the top — where the visual drama is greatest.

```{r step-2-order}
# See Lesson 1.2 (Aesthetics) — factor() with custom levels for meaningful y-axis order
country_order <- africa_measles %>%
  filter(year < 2001) %>%
  group_by(country) %>%
  summarise(avg_rate = mean(cases_per_million, na.rm = TRUE), .groups = "drop") %>%
  arrange(avg_rate) %>%
  pull(country)

africa_measles <- africa_measles %>%
  mutate(country = factor(country, levels = country_order))

# Top 5 highest-burden countries (pre-2001)
tail(country_order, 5)
```

### Step 3: Basic heatmap

```{r step-3-basic, fig.height=14}
# See Lesson 1.4 (Geometries) — geom_tile() creates a heatmap grid
ggplot(africa_measles, aes(x = year, y = country, fill = cases_per_million)) +
  geom_tile(colour = "white", linewidth = 0.4,
            width = 0.9, height = 0.9) +    # Tufte: tile spacing improves readability
  labs(title = "Step 3: Basic heatmap — default gradient hides the story") +
  theme_bka() +
  theme(panel.grid = element_blank(),
        axis.text.y = element_text(size = 9, face = "bold"))
```

The default gradient compresses everything because of extreme outliers. We need `scale_fill_gradientn()` with `oob = scales::squish` to handle the range.

### Step 4: Apply BKA gradient with outlier squishing

The data ranges from 0 to 33,000+ cases per million. Setting `limits = c(0, 5000)` captures 95% of the data; `oob = scales::squish` maps outliers beyond 5,000 to the maximum gradient color instead of turning them gray.

```{r step-4-gradient, fig.height=14}
# See Lesson 2.4 (Palettes) — colorRampPalette() + asymmetric split
measles_gradient <- c(
  colorRampPalette(c("#ACCBF9", "#005CB9", "#83BD00", "#F8A623"))(15),
  colorRampPalette(c("#F8A623", "#FA7650", "#E24A3F"))(5)
)

# See Lesson 2.4 (Palettes) — scale_fill_gradientn() for multi-stop gradients
ggplot(africa_measles, aes(x = year, y = country, fill = cases_per_million)) +
  geom_tile(colour = "white", linewidth = 0.4,
            width = 0.9, height = 0.9) +
  scale_fill_gradientn(
    colours = measles_gradient,
    limits = c(0, 5000),
    na.value = "#F5F5F5",
    oob = scales::squish,    # Values above 5000 get the max color, not gray
    name = "Cases per million"
  ) +
  labs(title = "Step 4: BKA gradient with squished outliers — pattern emerges") +
  theme_bka() +
  theme(panel.grid = element_blank(),
        axis.text.y = element_text(size = 9, face = "bold"))
```

Now the before/after pattern is visible: warm colors on the left (pre-2001), cooler colors on the right (post-2001).

### Step 5: Add the 2001 reference line and annotation

```{r step-5-annotation, fig.height=14}
ggplot(africa_measles, aes(x = year, y = country, fill = cases_per_million)) +
  geom_tile(colour = "white", linewidth = 0.4,
            width = 0.9, height = 0.9) +
  scale_fill_gradientn(
    colours = measles_gradient,
    limits = c(0, 5000),
    na.value = "#F5F5F5",
    oob = scales::squish,
    name = "Cases per million"
  ) +
  scale_x_continuous(expand = c(0, 0), breaks = seq(1980, 2020, by = 5)) +
  # See Lesson 3.2 (Annotations) — geom_vline() + annotate("text")
  geom_vline(xintercept = 2001, color = bka_colors$title_dark,
             linewidth = 1.2, alpha = 0.8) +    # Cairo: annotations anchor the narrative
  annotate("text", label = "WHO/UNICEF\nMeasles Initiative",
           x = 2001, y = length(country_order) + 0.5,
           hjust = -0.05, vjust = 1,
           size = 4, fontface = "bold", family = "Lato",
           color = bka_colors$title_dark) +
  coord_cartesian(clip = "off") +
  labs(title = "Step 5: The 2001 reference line anchors the before/after story") +
  theme_bka() +
  theme(panel.grid = element_blank(),
        axis.text.y = element_text(size = 9, face = "bold"))
```

In 2001, the WHO/UNICEF Measles & Rubella Initiative launched targeted supplementary immunization activities (SIAs) across Africa. The reference line shows where the color regime shifts from warm to cool for most countries.

### Step 6: Final polish — Minto title, legend, and BKA styling

```{r step-6-final, fig.height=14}
ggplot(africa_measles, aes(x = year, y = country, fill = cases_per_million)) +
  geom_tile(colour = "white", linewidth = 0.4,
            width = 0.9, height = 0.9) +
  scale_fill_gradientn(
    colours = measles_gradient,
    limits = c(0, 5000),
    breaks = seq(0, 5000, by = 1000),
    labels = c("0", "1K", "2K", "3K", "4K", "5K"),
    na.value = "#F5F5F5",
    name = "Reported cases per million population",
    oob = scales::squish,
    # See Lesson 2.5 (Legend Styling) — guide_colourbar() for continuous legends
    guide = guide_colourbar(
      barwidth = 15, barheight = 0.6,
      title.position = "top", title.hjust = 0.5,
      ticks = TRUE, nbin = 50
    )
  ) +
  scale_x_continuous(expand = c(0, 0), breaks = seq(1980, 2020, by = 5)) +
  geom_vline(xintercept = 2001, color = bka_colors$title_dark,
             linewidth = 1.2, alpha = 0.8) +
  annotate("text", label = "WHO/UNICEF\nMeasles Initiative",
           x = 2001, y = length(country_order) + 0.5,
           hjust = -0.05, vjust = 1,
           size = 4, fontface = "bold", family = "Lato",
           color = bka_colors$title_dark) +
  coord_cartesian(clip = "off") +
  # See Lesson 3.1 (Titles & Captions) — Minto-style insight title
  labs(
    title = "Mass immunization campaigns cut African measles cases by over 90%\nbut outbreaks persist where coverage gaps remain",
    subtitle = "Reported measles cases per million population by country, 1980-2023",
    x = NULL, y = NULL,
    caption = paste0(
      "Source: WHO Global Health Observatory (cases) & World Bank (population) | ",
      "Countries with 15+ years of reporting shown\n",
      "Original inspiration: r-graph-gallery.com/vaccination-heatmap.html"
    )
  ) +
  # See Lesson 2.2 (Custom Themes) — theme_bka() + heatmap-specific overrides
  theme_bka() +
  theme(
    panel.grid = element_blank(),    # Tufte: no gridlines for heatmaps
    axis.line = element_blank(),
    axis.text.y = element_text(size = 9, face = "bold", family = "Lato", hjust = 1),    # Few: readable with many rows
    axis.text.x = element_text(size = 8),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.title = element_text(size = 9, face = "bold",
                                color = bka_colors$title_dark),
    legend.text = element_text(size = 8, color = bka_colors$subtitle_gray),
    plot.title = element_text(lineheight = 1.15),
    plot.margin = margin(15, 15, 25, 15)
  )
```

## Three Heatmaps Compared

| Dimension | 01 — MNH (Synthetic) | 02 — US Measles (Original) | 03 — Africa Measles (This Lesson) |
|-----------|---------------------|---------------------------|----------------------------------|
| Data source | Local CSV | GitHub CSV | WHO + World Bank (pre-processed) |
| Rows | 8 regions | 51 states | 46 countries |
| Columns | 11 years | 84 years | 44 years |
| Metric | MMR per 100k | Cases per million | Cases per million |
| Reference line | 2020 HSTP II | 1963 vaccine | 2001 WHO/UNICEF Initiative |
| Missing data | None | Some (pre-reporting) | Common (surveillance gaps) |
| Outlier handling | None needed | None needed | `oob = scales::squish` |

## Data Notes

**What makes this chart different from the US version:**

1. **Missing data is real.** Gray tiles represent years where a country didn't report to WHO — this is itself informative, showing gaps in surveillance infrastructure.

2. **Outbreaks recur.** Unlike the US chart's clean post-vaccine silence, Africa shows periodic red spikes even after 2001. These represent real outbreaks when vaccination coverage dips below herd immunity thresholds (~95% for measles).

3. **The rollout was gradual.** Africa's EPI (Expanded Programme on Immunization) began incorporating measles vaccine in the late 1970s-1980s, but mass supplementary campaigns didn't accelerate until 2001. Some countries show early declines; others were late adopters.

4. **Per-capita normalization matters.** Without dividing by population, Nigeria's absolute case count would dominate the entire chart. Cases per million makes countries comparable. The data prep script (`../data/prep_africa_measles.R`) handles this join.

5. **Extreme outliers need squishing.** Some country-years exceed 30,000 cases per million. `oob = scales::squish` maps these to the gradient maximum instead of showing them as gray NA tiles.

### How the data was prepared

The CSV was generated by `../data/prep_africa_measles.R`, which:

1. Loads measles case counts from the WHO GHO API (`jsonlite::fromJSON()`)
2. Loads population from the World Bank API
3. Joins on country code + year
4. Computes `cases_per_million = cases / population * 1e6`
5. Maps ISO3 codes to country names via the WHO dimension lookup
6. Filters to countries with 15+ years of reporting

To refresh the data, re-run the prep script (requires internet and the `jsonlite` package).

## Quick Reference

| Technique | Function | Key Arguments | Lesson |
|-----------|----------|---------------|--------|
| Load local CSV | `read.csv()` | File path | — |
| Y-axis ordering | `factor(levels = ...)` | Custom sort order | 1.2 — Aesthetics |
| Heatmap grid | `geom_tile()` | `colour`, `linewidth`, `width`, `height` | 1.4 — Geometries |
| Multi-stop gradient | `scale_fill_gradientn()` | `colours`, `limits`, `oob` | 2.4 — Palettes |
| Squish outliers | `oob = scales::squish` | Inside `scale_fill_*()` | 2.4 — Palettes |
| Generate gradient | `colorRampPalette()` | Hex vectors, `(n)` | 2.4 — Palettes |
| Reference line | `geom_vline()` | `xintercept` | 3.2 — Annotations |
| Floating text | `annotate("text")` | `x`, `y`, `label` | 3.2 — Annotations |
| Colorbar legend | `guide_colourbar()` | `barwidth`, `barheight` | 2.5 — Legend Styling |
| Axis formatting | `scale_x_continuous()` | `expand`, `breaks` | 1.7 — Scales |
| Minto title | `labs(title = "...")` | Insight-first | 3.1 — Titles & Captions |
| Custom theme | `theme_bka()` + overrides | `panel.grid = element_blank()` | 2.2 — Custom Themes |

## Your Turn

Try filtering the heatmap to just the **"meningitis belt"** countries (Burkina Faso, Chad, Ethiopia, Mali, Niger, Nigeria, Senegal, Sudan) to see if the pattern differs for this high-burden sub-region.

```{r your-turn, eval=FALSE}
belt_countries <- c("Burkina Faso", "Chad", "Ethiopia", "Mali",
                    "Niger", "Nigeria", "Senegal", "___")

belt_measles <- africa_measles %>%
  filter(country %in% ___) %>%
  mutate(country = factor(country, levels = ___))

ggplot(belt_measles, aes(x = year, y = country, fill = cases_per_million)) +
  geom_tile(colour = "white", linewidth = ___,
            width = ___, height = ___) +
  scale_fill_gradientn(
    colours = measles_gradient,
    limits = c(___, ___),
    oob = scales::squish,
    name = "___",
    guide = guide_colourbar(
      barwidth = ___, barheight = ___,
      title.position = "top", title.hjust = 0.5
    )
  ) +
  scale_x_continuous(expand = c(0, 0), breaks = ___) +
  geom_vline(xintercept = ___, color = bka_colors$title_dark) +
  annotate("text", label = "___", x = ___, y = ___,
           hjust = -0.05, size = 3, family = "Lato",
           color = bka_colors$title_dark) +
  labs(
    title = "___",
    subtitle = "Meningitis belt countries — measles cases per million, 1980-2023",
    x = NULL, y = NULL,
    caption = "___"
  ) +
  theme_bka() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_blank(),
    legend.position = "bottom",
    legend.direction = "horizontal"
  )
```
