---
title: "Chapter 1: Why R for Management Consultants"
subtitle: "Motivation, installation, RStudio orientation, and first commands"
---

## Learning Objectives

::: {.learning-objectives}
After this chapter, you will be able to:

- Explain why **reproducibility** matters for consulting deliverables in maternal and newborn health (MNH)
- Navigate the four panes of the **RStudio IDE** (Source, Console, Environment, Files/Plots)
- Write and run your first **R commands** in the console and in an R script
- Install and load **packages** from CRAN
- Perform basic arithmetic with real-world MNH indicators (neonatal mortality rate, maternal mortality ratio)
:::

## Key Concepts

::: {.key-concept}
#### Reproducibility
The ability to re-run an analysis and get the same result. In consulting, this means your client can verify your numbers, your team can update a report when new data arrives, and you can trace every figure in a deliverable back to a specific line of code. Spreadsheet-based workflows (copy-paste, manual formatting) break reproducibility.
:::

::: {.key-concept .green}
#### RStudio IDE
An integrated development environment that wraps around R, providing a code editor with syntax highlighting, a console for interactive commands, an environment pane showing your data objects, and a files/plots/help pane. Think of R as the engine and RStudio as the dashboard.
:::

::: {.key-concept .teal}
#### R Scripts
Plain-text files (`.R` extension) that contain a sequence of R commands. Scripts are the basic unit of reproducible work — you write your analysis once, save it, and anyone can re-run it. Every consulting analysis should live in a script, not typed directly into the console.
:::

::: {.key-concept .orange}
#### Packages
Extensions that add new functions to R. The base installation handles arithmetic and basic statistics, but packages like `tidyverse`, `readxl`, and `ggplot2` provide the tools consultants actually need. You install a package once with `install.packages()` and load it each session with `library()`.
:::

## Lecture Slides

::: {.slide-embed}
<iframe src="../ch01-why-r/index.html"></iframe>
<div class="slide-embed-footer">
<a href="../ch01-why-r/index.html" target="_blank">Open slides full-screen &rarr;</a>
</div>
:::

## R Code Workshop

This workshop walks you through your very first R session — from basic arithmetic to creating variables that hold real MNH indicator data.

### Your First Calculations

```{r}
#| label: first-calculations
#| echo: true

# R works as a calculator
2 + 2
100 / 4
sqrt(144)

# Neonatal deaths per 1,000 live births (NMR formula)
# If a district had 45 neonatal deaths out of 3,200 live births:
45 / 3200 * 1000
```

The last line computes a neonatal mortality rate (NMR) of about 14.1 per 1,000 live births. In a spreadsheet you might type this into a cell — in R, you write it as a formula that anyone can read, verify, and re-run.

### Creating Variables

```{r}
#| label: creating-variables
#| echo: true

# Store values in named variables using the assignment operator <-
neonatal_deaths <- 45
live_births <- 3200

# Now compute NMR from the stored values
nmr <- neonatal_deaths / live_births * 1000
nmr

# Maternal mortality ratio (MMR): maternal deaths per 100,000 live births
maternal_deaths <- 12
mmr <- maternal_deaths / live_births * 100000
mmr
```

Variable names should be descriptive and use `snake_case` (lowercase with underscores). This makes your code self-documenting — a colleague reading `neonatal_deaths / live_births * 1000` immediately understands the calculation without needing a comment.

### Working with Multiple Values

```{r}
#| label: multiple-values
#| echo: true

# NMR for five countries (per 1,000 live births)
country_names <- c("Ethiopia", "Tanzania", "Malawi", "Kenya", "Rwanda")
nmr_values <- c(28.1, 20.3, 22.4, 18.6, 16.2)

# R operates element-wise — no loops needed
# Convert NMR to deaths per 10,000 births
nmr_per_10k <- nmr_values * 10
nmr_per_10k

# Which country has the highest NMR?
max(nmr_values)
country_names[which.max(nmr_values)]
```

The `c()` function combines values into a **vector** — R's fundamental data structure. You will work with vectors constantly. Notice how `nmr_values * 10` multiplied every element at once; this vectorized behavior is what makes R powerful for data analysis.

### Installing and Loading Packages

```{r}
#| label: installing-packages
#| echo: true

# Install the tidyverse (run this ONCE — it downloads from the internet)
# install.packages("tidyverse")

# Load it at the start of every script
library(tidyverse)

# Now you have access to dplyr, ggplot2, readr, and more
# Quick example: create a tibble (modern data frame)
mnh_data <- tibble(
  country = country_names,
  nmr = nmr_values
)
mnh_data
```

You only need to install a package once (like installing an app), but you must load it with `library()` every time you start a new R session (like opening the app).

## Key Takeaways

::: {.key-takeaway}
**R replaces spreadsheets** with a reproducible, scriptable workflow — essential for consulting teams that need to update analyses when new data arrives.

**RStudio is your workspace.** Learn the four panes: Source (write code), Console (run code), Environment (see your data), Files/Plots/Help (everything else).

**Variables store values, vectors store collections.** These are the building blocks for every analysis you will do in this course.
:::

## Looking Ahead

Next in **Chapter 2 — R Building Blocks**, we dive deeper into the data structures that power R:

- Vectors and their four main types (numeric, character, logical, factor)
- How R handles missing values (`NA`) — critical for real-world health data
- Writing your own custom functions for MNH indicator calculations
- Getting help with `?`, `help()`, and online resources
