---
title: "Chapter 2: R Building Blocks"
subtitle: "Vectors, data types, functions, and getting help"
---

## Learning Objectives

::: {.learning-objectives}
After this chapter, you will be able to:

- Create and manipulate **vectors** of numeric, character, logical, and factor data
- Check and convert between **data types** using `is.*()` and `as.*()` functions
- Explain how R handles **missing values** (`NA`) and why this matters for health data
- Write **custom functions** to encapsulate repeated calculations
- Use R's built-in **help system** (`?`, `help()`, `args()`) to learn about any function
:::

## Key Concepts

::: {.key-concept}
#### Vectors
The most fundamental data structure in R. A vector is an ordered collection of values — all of the same type. When you store NMR values for five countries, that is a numeric vector. When you store country names, that is a character vector. Even a single number in R is technically a vector of length 1.
:::

::: {.key-concept .green}
#### Data Types
Every value in R belongs to one of several types. The four you will use most: **numeric** (numbers like `28.1`), **character** (text like `"Ethiopia"`), **logical** (`TRUE` or `FALSE`), and **factor** (categorical variables with a fixed set of levels, like facility type). Mixing types in a vector triggers automatic coercion — R silently converts everything to the most general type.
:::

::: {.key-concept .red}
#### Missing Values (NA)
R uses `NA` to represent missing data — a facility that did not report, a country with no survey that year, a patient record with an unknown outcome. Most R functions will return `NA` if any input is `NA` unless you explicitly tell them to ignore missing values (e.g., `mean(x, na.rm = TRUE)`). Learning to handle `NA` properly is essential for working with real-world MNH datasets.
:::

::: {.key-concept .teal}
#### Functions
A named block of code that takes inputs (arguments), performs a calculation, and returns a result. R comes with thousands of built-in functions (`mean()`, `sum()`, `paste()`), and packages add thousands more. You can also write your own functions to encapsulate repeated calculations — like computing NMR or MMR from raw counts.
:::

::: {.key-concept .orange}
#### Getting Help
Every R function has a help page accessed with `?function_name` or `help(function_name)`. The help page shows the function's arguments, what it returns, and usage examples. When you are stuck, `?` is your first stop.
:::

## Lecture Slides

::: {.slide-embed}
<iframe src="../ch02-building-blocks/index.html"></iframe>
<div class="slide-embed-footer">
<a href="../ch02-building-blocks/index.html" target="_blank">Open slides full-screen &rarr;</a>
</div>
:::

## R Code Workshop

This workshop builds your fluency with R's core data types and shows you how to write functions that automate MNH indicator calculations.

### Creating and Inspecting Vectors

```{r}
#| label: creating-vectors
#| echo: true

# Numeric vector: NMR by country (per 1,000 live births)
nmr <- c(28.1, 20.3, 22.4, 18.6, 16.2)

# Character vector: country names
countries <- c("Ethiopia", "Tanzania", "Malawi", "Kenya", "Rwanda")

# Logical vector: does the country have NMR above 20?
high_nmr <- nmr > 20
high_nmr  # TRUE TRUE TRUE FALSE FALSE

# Factor vector: facility type (categorical)
facility_type <- factor(c("hospital", "health_center", "hospital",
                           "health_post", "health_center"),
                        levels = c("health_post", "health_center", "hospital"))
facility_type
```

Notice the logical vector: `nmr > 20` tests each element and returns `TRUE` or `FALSE`. This is how you will filter data later — by creating logical conditions that select rows.

### Type Checking and Conversion

```{r}
#| label: type-checking
#| echo: true

# Check types
is.numeric(nmr)       # TRUE
is.character(countries)  # TRUE
is.logical(high_nmr)  # TRUE

# What happens when types mix?
mixed <- c(1, 2, "three")
mixed  # "1" "2" "three" — everything becomes character

# Convert character to numeric (common with imported data)
reported_values <- c("120", "85", "NA", "200", "150")
as.numeric(reported_values)  # 120 85 NA 200 150

# Convert numeric to factor for categorical analysis
nmr_category <- cut(nmr, breaks = c(0, 15, 25, Inf),
                    labels = c("Low", "Medium", "High"))
nmr_category
```

Type coercion is one of the most common sources of bugs when importing data. A column that looks numeric in your CSV may be read as character if it contains a single non-numeric entry (like "N/A" or a footnote symbol). Always check your types after import.

### Handling Missing Values

```{r}
#| label: missing-values
#| echo: true

# Real-world data has gaps — some facilities did not report
deliveries <- c(340, 215, NA, 180, NA, 420, 310)

# Most functions propagate NA by default
mean(deliveries)      # NA — because R cannot know the true mean
sum(deliveries)       # NA

# Tell R to remove NAs before computing
mean(deliveries, na.rm = TRUE)   # 293
sum(deliveries, na.rm = TRUE)    # 1465

# Count and locate missing values
sum(is.na(deliveries))         # 2 missing values
which(is.na(deliveries))       # positions 3 and 5

# Percentage of data that is missing
mean(is.na(deliveries)) * 100  # 28.6% missing
```

In MNH data, missing values are not random — facilities that do not report often have the weakest infrastructure. Ignoring `NA` may bias your results. Always report how much data is missing and consider why it is missing before dropping it.

### Writing Custom Functions

```{r}
#| label: custom-functions
#| echo: true

# Function to calculate Neonatal Mortality Rate
calc_nmr <- function(neonatal_deaths, live_births) {
  nmr <- neonatal_deaths / live_births * 1000
  return(round(nmr, 1))
}

# Function to calculate Maternal Mortality Ratio
calc_mmr <- function(maternal_deaths, live_births) {
  mmr <- maternal_deaths / live_births * 100000
  return(round(mmr, 0))
}

# Use them
calc_nmr(45, 3200)     # 14.1 per 1,000 live births
calc_mmr(12, 3200)     # 375 per 100,000 live births

# Functions work on vectors too
district_neo_deaths <- c(45, 30, 62, 28, 51)
district_births <- c(3200, 2100, 4500, 1800, 3700)
calc_nmr(district_neo_deaths, district_births)
```

Writing functions eliminates copy-paste errors. Instead of re-typing the NMR formula in ten different places (and risking a typo in one), you define it once and call it everywhere. If the formula changes, you update one function.

### Using the Help System

```{r}
#| label: getting-help
#| echo: true

# Get help on any function
?mean
?round
?cut

# See a function's arguments
args(round)

# Search for functions by keyword
help.search("correlation")

# Example: discovering that mean() has an na.rm argument
?mean
# Usage: mean(x, trim = 0, na.rm = FALSE, ...)
# Setting na.rm = TRUE ignores missing values
```

The help system is your most reliable reference. Package vignettes (`vignette("dplyr")`) provide longer tutorials. For complex questions, searching "R [your question]" online will usually lead to a Stack Overflow answer.

## Key Takeaways

::: {.key-takeaway}
**Vectors are the foundation.** Every column of data you work with is a vector. Understanding their types (numeric, character, logical, factor) prevents subtle bugs.

**NA is not zero.** Missing data must be handled explicitly. Always check for `NA` values and decide whether to remove, impute, or flag them.

**Write functions for repeated calculations.** If you compute the same indicator more than once, wrap it in a function. This makes your code shorter, more readable, and less error-prone.
:::

## Looking Ahead

Next in **Chapter 3 — Importing & Exploring Data**, we move from hand-typed vectors to real datasets:

- Reading CSV and Excel files into R with `read_csv()` and `read_excel()`
- The `tibble` — a modern, friendlier version of the data frame
- Exploring data structure with `glimpse()`, `summary()`, and `str()`
- Systematic approaches to diagnosing missing values in imported data
