---
title: "R Environment Setup"
subtitle: "Installing R, RStudio, and the required packages for this course"
---

## Required Software

| Software | Version | Download |
|----------|---------|----------|
| **R** | 4.3 or later | [r-project.org](https://www.r-project.org/) |
| **RStudio** (recommended IDE) | Latest | [posit.co/downloads](https://posit.co/download/rstudio-desktop/) |
| **Quarto** (for Chapter 9) | 1.8 or later | [quarto.org](https://quarto.org/docs/get-started/) |

::: {.callout-box}
**Note:** The slide decks and this website use `eval: false` for all R code blocks, meaning R code is displayed but not executed during rendering. You do **not** need R installed to view the slides or website. However, to follow along with the exercises, you will need R and the packages below.
:::

## Package Installation

Run this once in your R console to install all packages used across the 9 chapters:

```{r}
#| label: install-all
#| eval: false

install.packages(c(
  "tidyverse",  # Includes dplyr, ggplot2, tidyr, readr, stringr, and more
  "readxl",     # Reading Excel files
  "scales",     # Axis formatting (percent, comma, dollar)
  "patchwork",  # Multi-panel chart composition
  "knitr"       # Tables in Quarto documents (kable)
))
```

## Packages by Chapter

| Package | Purpose | Chapters Used |
|---------|---------|---------------|
| (base R) | Vectors, data types, functions, basic stats | 1, 2, 8 |
| `readr` | `read_csv` for data import | 3, 4, 5, 6, 7, 8, 9 |
| `readxl` | `read_excel` for Excel files | 3 |
| `dplyr` | Data transformation (filter, mutate, summarise) | 4, 5, 6, 7, 8, 9 |
| `tidyr` | Reshaping data (pivot_longer, pivot_wider) | 5, 7 |
| `stringr` | String manipulation basics | 5 |
| `ggplot2` | Data visualization | 6, 7, 8, 9 |
| `scales` | Axis formatting (percent, comma) | 7, 8 |
| `patchwork` | Multi-panel chart layouts | 7 |
| `knitr` | `kable` for tables in Quarto | 9 |

## Verification Script

After installing, run this script to verify all packages load correctly:

```{r}
#| label: verify-packages
#| eval: false

# Verification: load each package and print version
packages <- c("readr", "readxl", "dplyr", "tidyr", "stringr",
              "ggplot2", "scales", "patchwork", "knitr")

for (pkg in packages) {
  if (requireNamespace(pkg, quietly = TRUE)) {
    cat(sprintf("✓ %s %s\n", pkg, packageVersion(pkg)))
  } else {
    cat(sprintf("✗ %s — NOT INSTALLED\n", pkg))
  }
}
```

Expected output (version numbers may vary):

```
✓ readr 2.1.5
✓ readxl 1.4.3
✓ dplyr 1.1.4
✓ tidyr 1.3.1
✓ stringr 1.5.1
✓ ggplot2 3.5.1
✓ scales 1.3.0
✓ patchwork 1.2.0
✓ knitr 1.45
```

## Troubleshooting

### tidyverse installation issues on Windows

If `install.packages("tidyverse")` fails:

1. Install [Rtools](https://cran.r-project.org/bin/windows/Rtools/) (match the version to your R version)
2. Restart R and try `install.packages("tidyverse")` again

### Package installation behind a proxy

If you are on a corporate network with a proxy, set it before installing:

```{r}
#| label: proxy-setup
#| eval: false

Sys.setenv(http_proxy = "http://your-proxy:port")
Sys.setenv(https_proxy = "http://your-proxy:port")
install.packages("tidyverse")
```

### readxl on older R versions

If `readxl` installation fails, ensure your R version is 4.3 or later. Update R from [r-project.org](https://www.r-project.org/).

## Consistent MNH Variables

These variable names are used consistently across chapters in the R code examples:

| Variable | Description | Example Values | R Name |
|----------|-------------|----------------|--------|
| Facility ID | Unique identifier | F001, F002, ... | `facility_id` |
| Country | Country name | Ethiopia, Tanzania, Malawi, Mozambique, Kenya | `country` |
| Facility Type | Level of care | Hospital, Health Center, Health Post | `facility_type` |
| Staffing Level | Staff adequacy | Adequate, Low | `staffing_level` |
| CPAP Available | Equipment status | TRUE / FALSE | `has_cpap` |
| Delivery Volume | Monthly deliveries | 15--200 | `delivery_volume` |
| NMR | Neonatal mortality rate | Per 1,000 live births | `nmr` |
| MMR | Maternal mortality ratio | Per 100,000 live births | `mmr` |
| ANC Coverage | Antenatal care coverage | 0--100% | `anc_coverage` |
| Training | Training completion | TRUE / FALSE | `training_completed` |
